---
title: "First Trends Dry Run"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r message = FALSE}
library(tidyverse)
library(brms)
library(posterior)
```


```{r}
NHM_ForFitting <- read_csv(  '../data/FilteredSizeAndLoc_Seasons_DRYRUN.csv')%>%
  select(Species= `Taxon Name` , 
         Year = year, 
         Sex = gender, 
         wingspan_mm, 
         Brood, 
         decimalLatitude ,
         Develop_Season =TempDEVELOP ,
         Develop_Temp= SeasonTT_Develop, 
         Laying_Temp = SeasonTT_Lay) %>%
  mutate( Sex = ifelse( Sex == 'N/A', 'Both', Sex))
```

### Correlations between year and temperature


Looks like there is a trend, but it is not the explanatory
However note there are  trends in location wrapped up in all of this. Over time, more records from more southerly locations. 

```{r}
NHM_ForFitting %>%
  ggplot( aes( x = Year, y = Develop_Temp))+
  geom_point( aes(col =decimalLatitude))+
  geom_smooth(method = 'lm')+
  facet_wrap(~Develop_Season)

NHM_ForFitting %>%
  ggplot( aes( x = Year, y = decimalLatitude))+
  geom_point( aes(col =Develop_Temp))+
  geom_smooth(method = 'lm')+
  facet_wrap(~Develop_Season)

NHM_ForFitting %>%
  filter(Develop_Season == 'SPRING') %>%
  lm(data =., Develop_Temp ~ decimalLatitude +Year)%>%
  summary()

```

### Fitting models

```{r}
## Cycle through each species to fit models 

Batches<- count(NHM_ForFitting,
                Species, Brood ,Sex)%>%
  arrange( n)

## 200 potential models to fit:

Batches$n %>% log10 %>% plot # almost perfect log-normal plot

sum(Batches$n >100) # 155 have over 100, 

Batches %>%
  filter( n >100 ) -> TableOfModels



Batches %>%
  pivot_wider(values_from = 'n',  names_from = 'Sex', values_fill = 0) %>%
  mutate( Perc_Undetermined = 100* Both / ( male+Both+female)  ) -> PercentageUndetermined

PercentageUndetermined %>%
  ggplot( aes( x =Perc_Undetermined, fill = Brood))+
  geom_histogram()

## percentage undtermined is pretty consistence between broods 
PercentageUndetermined%>%
  ggplot( aes( y =Perc_Undetermined, x = Species,col = Brood))+
  geom_point()

NHM_ForFitting%>%
  group_by(Year) %>%
  summarise(FracUndetermined = mean(Sex=='Both')) %>%
  ggplot(aes(y= FracUndetermined, x = Year ))+
  geom_point()

### For spotting weirdnesses
NHM_ForFitting %>%
  group_by(Species, Brood) %>%
  summarise( signif(mean(Year),1), 
             min(Year), 
             quantile(Year, 0.1), 
             quantile(Year, 0.9),
             max(Year), 
             n = n())%>%
  mutate(EightyQuantileRange = `quantile(Year, 0.9)`-`quantile(Year, 0.1)` ) %>%
  arrange( EightyQuantileRange)%>%
  arrange(Species            ) -> NHM_Brood_data_spread

NHM_Brood_data_spread %>%
  left_join(PercentageUndetermined) %>%
write_csv('../data/NHM_Brood_data_spread.csv')

```


```{r}
Set = paste0('Sp.', TableOfModels$Species,
             '_Bro.',TableOfModels$Brood,
             '_Sex.',TableOfModels$Sex)


```

```{r eval = FALSE}
### Define and compile model
CoreModel <-   NHM_ForFitting %>%
    filter(Species == TableOfModels$Species[1],
           Brood   == TableOfModels$Brood[1]  ,
           Sex     == TableOfModels$Sex[1]) %>%
    brm( wingspan_mm ~ Year + Develop_Temp  + decimalLatitude,
         data = ., refresh = 0)

## Refit model to each set of data, saving each time
for( i in 1:nrow( TableOfModels)){
  print(Set[i])
  NHM_ForFitting %>%
    filter(Species == TableOfModels$Species[i],
           Brood   == TableOfModels$Brood[i]  ,
           Sex     == TableOfModels$Sex[i]) -> batchDATA
  update( CoreModel, newdata = batchDATA, refresh = 0) -> MODEL_FIT
  save(MODEL_FIT, file = paste('../NHM_modelfits2/',Set[i]  ))
}

```

### Collating model outputs into a df

```{r eval = FALSE}
map_df(1:nrow( TableOfModels), function(i){
  load(file = paste('../NHM_modelfits2/',Set[i]  )) 
  MODEL_FIT %>%
    as_draws_array %>%
    summarise_draws -> Fit_df
  Fit_df$Species = TableOfModels$Species[i]
  Fit_df$Brood = TableOfModels$Brood[i] 
  Fit_df$Sex = TableOfModels$Sex[i] 
  
  return(Fit_df)
})-> Posterior_Summaries

write_csv(Posterior_Summaries, 'Posterior_Summaries_dryrun2.csv')

```


## Step 2 - Does ecologucal specificity predict the trends?

Does ecological specificity determine trends?


```{r}
Posterior_Summaries<- read_csv('Posterior_Summaries_dryrun2.csv')


Posterior_Summaries %>%
  filter( variable %in% c( 'b_Develop_Temp',
                           'b_Year',
                           'b_decimalLatitude' ) )%>%
  ggplot( aes( y =  paste( Species,Brood,Sex ), 
               x = mean))+
  geom_point()+
  geom_errorbarh(aes( xmin = q5, xmax = q95))+
  geom_vline(xintercept = 0)+
    facet_wrap(~variable, scales = 'free_x')

## Join to species predictors

## fit including uncertainty in coefficent. 


ggsave(filename = 'AllCombos_Trends_DryRun.pdf', width = 8, height =20)

```












```{R}
#sessionInfo()

```