---
title: "Lepidoptera_code"
author: "Elsa Heywood"
date: "2024-03-18"
output:
  word_document: default
  html_document: default
---


Checks that my dataset isn't biased

Check the coverage of my data

```{r}
library(ggplot2)
ggplot(Project_data_copynew, aes(x = Year)) +
  geom_bar(fill = "#2895F5", width = 0.7) +  # Set the width of bars to 0.7
  labs(title = "Coverage of Specimens Across Years", x = "Year", y = "Count") +
  theme_minimal()
```


Add latitude as a continuous variable to dataset

```{r}
midpoints <- c(1.5, 2.5, 3.5, 4.5, 5.5, 6.5, 7.5, 8.5, 9.5)
Project_data_copynew$Latitude_Continuous <- midpoints[Project_data_copynew$Lattitude_Group]
```

Check species shifting latitude over time isn't causing effects

```{r}
library(ggplot2)
library(colorspace)

# Generate a color palette with 29 distinct colors
species_palette <- rainbow_hcl(29)

# Plot Latitude_Continuous against Year, with different lines for each Species
ggplot(Project_data_copynew, aes(x = Year, y = Latitude_Continuous, color = Species)) +
  geom_point() +
  scale_color_manual(values = species_palette) +
  labs(x = "Year", y = "Latitude Group", color = "Species") +
  theme_minimal() +
  facet_wrap(~Species)+
  geom_smooth(col='black')+guides(col='none')
```


Z transform my data by species

```{r}
library(dplyr)
# Define a function for log transformation
scale_this <- function(x){
  (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE)
}

unscale_this <- function(x, sd_z, mean_z){
  (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE)
}

# Applying custom scaling function within each group
Project_data_copynew <- 
  Project_data_copynew %>%
  group_by(Species) %>%
  mutate(Z_Forewing_Area = scale_this(Forewing_Area),
         Z_Wing_Length = scale_this(Wing_Length),
         Z_Abdomen_Length = scale_this(Abdomen_Length),
         Z_Abdomen_Width = scale_this(Abdomen_Width))
```


Examine the overall changes over time
Build the models

```{r}
area_model <- lm(Z_Forewing_Area ~ Year, data = Project_data_copynew)
abwidth_model <- lm(Z_Abdomen_Width ~ Year, data = Project_data_copynew)
wing_length_model <- lm(Z_Wing_Length ~ Year, data = Project_data_copynew)
ablength_model <- lm(Z_Abdomen_Length ~ Year, data = Project_data_copynew)
```

Look at results

```{r}
summary(area_model)
summary(abwidth_model)
summary(wing_length_model)
summary(ablength_model)
```

Visualise wing area over time

```{r}
#load ggplot
library(ggplot2)
library(broom)

# Extract coefficients and p-value
area_model_summary <- tidy(area_model)

# Filter the p-value for the specific coefficient you are interested in
# Let's say you want the p-value for the coefficient of Year
p_value <- area_model_summary$p.value[area_model_summary$term == "Year"]

# Create the plot
ggplot(Project_data_copynew, aes(x = Year, y = Z_Forewing_Area, color = Species)) +
  geom_point() +  # Scatter plot
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x, color = "black") +  # Overall regression line
  annotate("text", x = max(Project_data_copynew$Year),
           y = max(Project_data_copynew$Z_Forewing_Area),
           label = paste("p-value =", formatC(p_value, digits = 2)),
           hjust = 1, vjust = 0.1, size = 4, color = "black") +  # Add p-value text
  labs(title = " Z Forewing Area Changes Over Time by Species",
       x = "Year",
       y = "Z Forewing Area",
       color = "Species") +
  theme_minimal() 
```
Visualise abdomen width over time

```{r}
#load ggplot
library(ggplot2)
library(broom)

# Extract coefficients and p-value
abwidth_model_summary <- tidy(abwidth_model)

# Filter the p-value for the specific coefficient you are interested in
# Let's say you want the p-value for the coefficient of Year
p_value <- abwidth_model_summary$p.value[abwidth_model_summary$term == "Year"]

# Create the plot
ggplot(Project_data_copynew, aes(x = Year, y = Z_Abdomen_Width, color = Species)) +
 geom_point() +  # Scatter plot
 geom_smooth(method = "lm", se = FALSE, formula = y ~ x, color = "black") +  # Overall regression line
  annotate("text", x = max(Project_data_copynew$Year),
           y = max(Project_data_copynew$Z_Abdomen_Width),
           label = paste("p-value =", formatC(p_value, digits = 2)),
           hjust = 1, vjust = 0.1, size = 4, color = "black") +  # Add p-value text
  labs(title = " Z Abdomen Width Changes Over Time by Species",
       x = "Year",
       y = "Z Abdomen Width",
       color = "Species") +
  theme_minimal()
```
Visualise what individual species are doing over time

```{r}
#forewing area
Project_data_copynew %>%
  ggplot(aes(y = Forewing_Area, x = Year)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add regression line
  facet_wrap(~Species, scales = 'free_y')

#abdomen width
Project_data_copynew %>%
  ggplot(aes(y = Abdomen_Width, x = Year)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add regression line
  facet_wrap(~Species, scales = 'free_y')
```
Looking at the rates of change in each species across the time period

```{r}
library(dplyr)
library(broom)
library(tidyr)

# Define a function to calculate the overall rate of change for each species
calculate_overall_rate_per_species <- function(Project_data_copynew, Forewing_Area) {
  species_data <- split(Project_data_copynew, Project_data_copynew$Species)
  overall_rates <- sapply(species_data, function(sub_data) {
    model <- lm(substitute(Forewing_Area ~ Year), data = sub_data)
    coef(model)["Year"]
  })
  overall_rates_df <- data.frame(Species = names(overall_rates), Overall_Rate = overall_rates, row.names = NULL)
  return(overall_rates_df)
}

# Calculate the overall rate of change for each species for Forewing_Area using your data
forewing_overall_rate_per_species <- calculate_overall_rate_per_species(Project_data_copynew, Project_data_copynew$Forewing_Area)

# Print the table
print(forewing_overall_rate_per_species)

##same for abdomen width
calculate_overall_rate_per_species_ab <- function(Project_data_copynew, Abdomen_Width) {
  species_data_ab <- split(Project_data_copynew, Project_data_copynew$Species)
  overall_rates_ab <- sapply(species_data_ab, function(sub_data) {
    model <- lm(substitute(Abdomen_Width ~ Year), data = sub_data)
    coef(model)["Year"]
  })
  overall_rates_df_ab <- data.frame(Species = names(overall_rates_ab), Overall_Rate = overall_rates_ab, row.names = NULL)
  return(overall_rates_df_ab)
}

# Calculate the overall rate of change for each species for Abdomen width using your data
ab_overall_rate_per_species <- calculate_overall_rate_per_species_ab(Project_data_copynew, Project_data_copynew$Abdomen_Width)

# Print the table
print(ab_overall_rate_per_species)
```



What factors explain these changes? main model

multiple regression on forewing area and abdomen width

```{r}
multiple_model <- lm(Z_Forewing_Area ~Year + Temperature_Before + Ecology + Latitude_Continuous + Family + Family*Year + Ecology*Year + Ecology*Temperature_Before, data = Project_data_copynew)

multiple_model_ab <- lm(Z_Abdomen_Width ~Year + Temperature_Before + Ecology + Latitude_Continuous + Family + Family*Year + Ecology*Year + Ecology*Temperature_Before, data = Project_data_copynew)
```

Examine results

```{r}
summary(multiple_model)
anova(multiple_model)

summary(multiple_model_ab)
anova(multiple_model_ab)
```

Ecology*Year interaction so lets look at the trends within ecology

Subset the data by ecology

```{r}
# Create separate datasets for each group
group_SS <- subset(Project_data_copynew, Species %in% c("swallowtail", "white admiral", "adonis blue", "large heath", "silver spotted skipper", "silver washed fritillary", "small blue"))
group_GS <- subset(Project_data_copynew, Species %in% c("brimstone", "dark green fritillary", "peacock", "purple hairstreak", "small copper", "small skipper", "small tortoiseshell", "large white"))
group_GG <- subset(Project_data_copynew, Species %in% c("brown argus", "common blue", "comma", "gatekeeper", "grizzled skipper", "holly blue", "meadow brown", "orange tip", "speckled wood", "wall", "wood white", "dingy skipper", "marbled white"))
```


Multiple regression wing area within ecology

```{r}
#multiple regression and anova for forewing changes within ecological group
model_SS <- lm(Z_Forewing_Area ~ Temperature_Before + Year + Species + Year*Species + Temperature_Before*Species + Latitude_Continuous, data = group_SS)
summary(model_SS)
anova(model_SS)
model_GS <- lm(Z_Forewing_Area ~ Temperature_Before + Year + Species + Year*Species + Temperature_Before*Species + Latitude_Continuous, data = group_GS)
summary(model_GS)
anova(model_GS)
model_GG <- lm(Z_Forewing_Area ~ Temperature_Before + Year + Species + Year*Species+ Temperature_Before*Species + Latitude_Continuous, data = group_GG)
summary(model_GG)
anova(model_GG)
```

Multiple regression abdomen width within ecology

```{r}
#multiple regression and anova for abdomen changes within ecological group
model_ab_SS <- lm(Z_Abdomen_Width ~ Temperature_Before + Year + Species + Year*Species+ Temperature_Before*Species + Latitude_Continuous, data = group_SS)
summary(model_ab_SS)
anova(model_ab_SS)
model_ab_GS <- lm(Z_Abdomen_Width ~ Temperature_Before + Year + Species + Year*Species + Temperature_Before*Species + Latitude_Continuous, data = group_GS)
summary(model_ab_GS)
anova(model_ab_GS)
model_ab_GG <- lm(Z_Abdomen_Width ~ Temperature_Before + Year + Species + Year*Species+ Temperature_Before*Species + Latitude_Continuous, data = group_GG)
summary(model_ab_GG)
anova(model_ab_GG)
```
Visualise wing and abdomen trends

```{r}
# Plot the forewing area over time
ggplot(group_SS, aes(x = Year, y = Z_Forewing_Area)) +
  geom_point(color = "#EFA028") +  # Scatter plot
  labs(title = "Forewing area over time in habitat and food plant specialists",
       x = "Year",
       y = "Forewing area") +
  theme_minimal() +  # Optional: Adjust the theme if needed
  # Calculate linear regression
  geom_abline(slope = coef(lm(Z_Forewing_Area ~ Year, data = group_SS))[2],
              intercept = coef(lm(Z_Forewing_Area ~ Year, data = group_SS))[1],
              color = "black") +
  annotate(
    "text",
    x = max(group_SS$Year),
    y = max(group_SS$Z_Forewing_Area),
    label = paste("p =", formatC(0.74776, digits = 3)),
    hjust = 1,
    vjust = 1,
    color = "black"
  )

# Plot the forewing area over time
ggplot(group_GS, aes(x = Year, y = Z_Forewing_Area)) +
  geom_point(color = "#EFA028") +  # Scatter plot
  labs(title = "Forewing area over time in habitat generalists and food plant specialists",
       x = "Year",
       y = "Forewing area") +
  theme_minimal() +  # Optional: Adjust the theme if needed
  # Calculate linear regression
  geom_abline(slope = coef(lm(Z_Forewing_Area ~ Year, data = group_GS))[2],
              intercept = coef(lm(Z_Forewing_Area ~ Year, data = group_GS))[1],
              color = "black") +
  annotate(
    "text",
    x = max(group_GS$Year),
    y = max(group_GS$Z_Forewing_Area),
    label = paste("p =", formatC(0.0001948, digits = 3)),
    hjust = 1,
    vjust = 1,
    color = "black"
  )

# Plot the forewing area over time
ggplot(group_GG, aes(x = Year, y = Z_Forewing_Area)) +
  geom_point(color = "#EFA028") +  # Scatter plot
  labs(title = "Forewing area over time in habitat and food plant generalists",
       x = "Year",
       y = "Forewing area") +
  theme_minimal() +  # Optional: Adjust the theme if needed
  # Calculate linear regression
  geom_abline(slope = coef(lm(Z_Forewing_Area ~ Year, data = group_GG))[2],
              intercept = coef(lm(Z_Forewing_Area ~ Year, data = group_GG))[1],
              color = "black") +
  annotate(
    "text",
    x = max(group_GG$Year),
    y = max(group_GG$Z_Forewing_Area),
    label = paste("p =", formatC(0.0003496, digits = 3)),
    hjust = 1,
    vjust = 1,
    color = "black"
  )

# Plot the abdomen width over time
ggplot(group_SS, aes(x = Year, y = Z_Abdomen_Width)) +
  geom_point(color = "#2895F5") +  # Scatter plot
  labs(title = "Abdomen width over time in habitat and food plant specialists",
       x = "Year",
       y = "Abdomen width") +
  theme_minimal() +  # Optional: Adjust the theme if needed
  # Calculate linear regression
  geom_abline(slope = coef(lm(Z_Abdomen_Width ~ Year, data = group_SS))[2],
              intercept = coef(lm(Z_Abdomen_Width ~ Year, data = group_SS))[1],
              color = "black") +
  annotate(
    "text",
    x = max(group_SS$Year),
    y = max(group_SS$Z_Abdomen_Width),
    label = paste("p =", formatC(0.97486 , digits = 3)),
    hjust = 1,
    vjust = 1,
    color = "black"
  )

# Plot the abdomen width over time
ggplot(group_GS, aes(x = Year, y = Z_Abdomen_Width)) +
  geom_point(color = "#2895F5") +  # Scatter plot
  labs(title = "Abdomen width over time in habitat generalists and food plant specialists",
       x = "Year",
       y = "Abdomen width") +
  theme_minimal() +  # Optional: Adjust the theme if needed
  # Calculate linear regression
  geom_abline(slope = coef(lm(Z_Abdomen_Width ~ Year, data = group_GS))[2],
              intercept = coef(lm(Z_Abdomen_Width ~ Year, data = group_GS))[1],
              color = "black") +
  annotate(
    "text",
    x = max(group_GS$Year),
    y = max(group_GS$Z_Abdomen_Width),
    label = paste("p =", formatC(7.253e-06 , digits = 3)),
    hjust = 1,
    vjust = 1,
    color = "black"
  )

# Plot the abdomen width over time
ggplot(group_GG, aes(x = Year, y = Z_Abdomen_Width)) +
  geom_point(color = "#2895F5") +  # Scatter plot
  labs(title = "Abdomen width over time in habitat and food plant generalists",
       x = "Year",
       y = "Abdomen width") +
  theme_minimal() +  # Optional: Adjust the theme if needed
  # Calculate linear regression
  geom_abline(slope = coef(lm(Z_Abdomen_Width ~ Year, data = group_GG))[2],
              intercept = coef(lm(Z_Abdomen_Width ~ Year, data = group_GG))[1],
              color = "black") +
  annotate(
    "text",
    x = max(group_GG$Year),
    y = max(group_GG$Z_Abdomen_Width),
    label = paste("p =", formatC(0.0007981 , digits = 3)),
    hjust = 1,
    vjust = 1,
    color = "black"
  )
```





Look at trends within family over time

```{r}
# Plot Z_Forewing_Area over time, colored by Family with trend lines only
ggplot(Project_data_copynew, aes(x = Year, y = Z_Forewing_Area, color = Family)) +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear trend lines
  scale_color_manual(values = c("#F5AA0D", "#AA51DC", "#045A8A", "#0E6711", "#67070F")) +  # You can specify colors as per your preference
  labs(x = "Year", y = "Forewing Area", title = "Forewing Area Over Time by Family")

ggplot(Project_data_copynew, aes(x = Year, y = Z_Abdomen_Width, color = Family)) +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear trend lines
  scale_color_manual(values = c("#F5AA0D", "#AA51DC", "#045A8A", "#0E6711", "#67070F")) +  # You can specify colors as per your preference
  labs(x = "Year", y = "Abdomen Width", title = "Abdomen Width Over Time by Family")

```

