---
title: "Quick Spin through NHM data"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using data from: Wilson et al https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13844


Sourced from: 
https://zenodo.org/records/5759759

and
https://zenodo.org/records/6066070


```{r message =FALSE}
library( tidyverse)
library(readxl)
library(brms)
library(posterior)

Project_data_copynew <- read_excel("../data/Project_data_copynew.xlsx", sheet = "Sheet1") 
```


## Loading in measurments

```{r}
## Different groups saved in seperate sheets

# Sheets<- excel_sheets('../data/NHM/Wilson_et_al_MEE_data.xlsx')
# 
# map_df()
# 
# read_xlsx('../data/NHM/Wilson_et_al_MEE_data.xlsx', 
#             )

WilsonSH3<- read_xlsx('../data/NHM/Wilson_et_al_MEE_data.xlsx', sheet = 3)
Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')
```

### Downloading locations

Gordon L J Paterson; Sara Albuquerque; Vladimir Blagoderov; Stephen J Brooks et al. (2016). iCollections [Data set]. Natural History Museum. https://doi.org/10.5519/0038559

 *DOI has been created for this download request: https://doi.org/10.5519/qd.tqng8ri6. Please ensure you reference this DOI when citing these data.
For more information, follow the DOI link.*


```{r eval = FALSE}
Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')

NHM_data <- read_csv("C:/Users/pemb4504/Documents/7ba5ee8575924a906078531b9212aec880045725/resource.csv")

colnames(NHM_data)

NHM_data %>%
  select( "_id","basisOfRecord","catalogNumber",
          "class","collectionCode",
          "continent","coordinateUncertaintyInMeters", "country",                 
          "created","day","decimalLatitude","decimalLongitude" ,            
          "determinationFiledAs","determinationNames",  
          "family", "gbifID"    ,                   
          "gbifIssue", "genus",
          "higherClassification","month",      
          "order","project" ,                     
          "recordedBy" ,                 
          "scientificName",               
          "verbatimLatitude","verbatimLongitude","year" )      -> NHM_data_firstcut  

ifelse(sapply(NHM_data_firstcut, function(x)all(is.na(x))) == TRUE, "Y","N")

NHM_data_firstcut %>%
  mutate( `specimen number` = parse_number(catalogNumber))  %>%
  select( catalogNumber,  `specimen number`,determinationNames ,
          decimalLatitude,decimalLongitude, year, month )-> Just_Names_Loc_and_Numbers

write_csv(Just_Names_Loc_and_Numbers, '../data/Just_Names_Loc_and_Numbers.csv')
```


```{r warning=FALSE}
Just_Names_Loc_and_Numbers <- read_csv( '../data/Just_Names_Loc_and_Numbers.csv')

left_join(Sizes, Just_Names_Loc_and_Numbers) -> Joined_sizes_and_Locs

 
 Joined_sizes_and_Locs%>%
  rename(wingspan_mm =`wing_span (mm)` ) %>%
  filter(!is.na( decimalLatitude ),  ## Excluding those without locations 
         wingspan_mm !=0, ## Exclude those without measurements  
         decimalLatitude <54, ## Exclude Northerly samples
         decimalLongitude > -5.5, ## Exclude Irish samples 
         wingspan_mm < 100) %>%   ## Exclude implausibly high measurements 
  group_by(`Taxon Name`)%>%
  add_count(name = 'Total_Obs') %>%
  filter( Total_Obs>50) %>%## Exclude species with very few observations
  ungroup -> FilteredSizeAndLoc
 
### Add Species filterring here 
 
write_csv(FilteredSizeAndLoc,  '../data/FilteredSizeAndLoc.csv') 
 
```
 
 
```{r}
 

## Map of spread of samples: 
Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
  ggplot(aes( x = decimalLongitude, y = decimalLatitude, col = year))+
   geom_point()

Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
  count(Family, sort = TRUE)

#  Family      n
#Lycaenidae   70706  # Gossamer-winged butterflies
#Nymphalidae  64543  # Brush-footed butterflies
#Pieridae     18779
#Hesperiidae   8036  # Skipper butteflies
#Riodinidae    1526  # Metalmark butterflies
#Papilionidae   730   # Swallowtail butterlfies

## Many of the single observations are tropical, some are very rare vagrants (e.g. Glaucopsyche alexis) 
Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
  count(`Taxon Name`)%>%
  arrange( n)

## Distribution of measurements
Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
   ggplot(aes( y = `wing_span (mm)`, x = year))+
   geom_point()

```

# Temperature effects

```{r}
##### Check these numbers and their meaning!!
Temps<- read_csv( '../data/NHM/temperatures_SEngland.csv') %>%
  mutate( SummerTemp_YearBefore = lag(`Summer temperature` ))%>%
  rename(year = Year)
```

## Cleaning up and joining
```{r}



FilteredSizeAndLoc$year %>% hist

FilteredSizeAndLoc%>%
  left_join(Temps, by = 'year')-> Filt_BF_step1


```


### Excluding post-July

Recording Months are coded in a messy way:

 month     n
 3 - 4     1
 5 - 9     1
 8 - 7     1
 6 - 8     2
11 - 12     4
 4 - 5     7
 8 - 9     9
 3 - 6    12
 6 - 9    21
    12    31
     1    38
     2    62
    11    75
 6 - 7   141
 5 - 6   158
     3   176
 7 - 8   262
    10  1194
     4  1570
     9  9052
     5 15160
     6 24982
  <NA> 33704
     8 36839
     7 40818

```{r}
Filt_BF_step1 %>%
  mutate( Month_av = case_when(month == '3 - 4'   ~ 3.5  ,
                               month == '5 - 9'   ~ 7  ,
                               month == '8 - 7'   ~ 7.5  ,
                               month == '6 - 8'   ~ 7  ,
                               month == '11 - 12' ~ 11.5  ,
                               month == '4 - 5'   ~ 4.5  ,
                               month == '8 - 9'   ~ 8.5  ,
                               month == '3 - 6'   ~ 4.5  ,
                               month == '6 - 9'   ~ 7.5   ,
                               month == '6 - 7'   ~ 6.5,
                               month == '5 - 6'   ~ 5.5  ,
                               month == '7 - 8'   ~ 7.5   , 
                              .default = as.numeric(month)),
            FirstFlight = Month_av < 7 )->  Filt_BF



### majority of observations from later half of year
### -> probably need to only filer our bivoltine species, which will need collation
Filt_BF%>%
count(FirstFlight) 

Filt_BF %>%
  ggplot(aes( x = Month_av))+
  geom_histogram()+
  facet_wrap(~`Taxon Name`, scales = 'free_y')+
  geom_vline(xintercept = 7, col = 'red')


```



## Correlation in wing measurements

All are highly correlated except 'wing_shoulder'. Will just use 'wingspan'

```{r}
Filt_BF %>%
  select(ends_with('(mm)')) %>%
  cor
```

## Plots

```{r fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
Filt_BF %>%
  ggplot(aes( x = year, y=  wingspan_mm))+
  geom_point()+
  geom_smooth(method = 'lm')+
  facet_wrap(~`Taxon Name`, scales = 'free_y')+
  ggtitle('some trends in size of time?')


Filt_BF %>%
  ggplot(aes( x = SummerTemp_YearBefore, y=  wingspan_mm))+
  geom_point()+
  geom_smooth(method = 'lm')+
  facet_wrap(~`Taxon Name`, scales = 'free_y')+
  ggtitle('not a lot of impact of year-before temperature?')


Filt_BF %>%
  ggplot(aes( x = `Spring temperature`, y=  wingspan_mm))+
  geom_point()+
  geom_smooth(method = 'lm')+
  facet_wrap(~`Taxon Name`, scales = 'free_y')+
  ggtitle('some trends in size?')

```

# Step 1 - What is the trend of each species with temperature?

## Further cleaning

Excluding:
 - observations from second generations in bivoltine species (After July (?))
 - migratory and extinct species
 - observations outside core area

```{r}

MigratorySpecies_To_Exclude  <- c('Vanessa cardui')  ### TODO EXTEND!
ExtinctSpecies_To_Exclude  <- c()  ### TODO EXTEND!


Species_list<- c('Aglais urticae', 'Inachis io')   ## Trial species
  ### Sp = 'Aglais urticae'

Filt_BF %>%
  count(`Taxon Name`) %>% 
  as.data.frame()

```


[[This now moved to First Trends Dry Run doc]]



# Session info

```{r}
sessionInfo()
```





