---
title: "Temperature maps"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



County shape files with temperature records at seasonal level, for each year. 

```{r message=FALSE, warning=FALSE}
library(sf) # for polygons
library(tidyverse) # for general manipulation
library(terra) # for rasters
library(tidyterra) # for plotting rasters 
library(exactextractr)  ## for faster avaerging of raster polygon intersections
```


## Read in County shape files

Unitary Authorities:

https://www.data.gov.uk/dataset/85228aec-fe0e-49bf-9455-df000d61e731/counties-and-unitary-authorities-may-2023-boundaries-uk-bgc

Vice Counties 

From https://www.brc.ac.uk/article/british-vice-counties / https://github.com/BiologicalRecordsCentre/vice-counties/tree/master

```{r message=FALSE}
LegalCounties<- st_read('../data/Counties_and_Unitary_Authorities_May_2023_UK_BGC_9070630924157476989/CTYUA_MAY_2023_UK_BGC.shp')

LegalCounties %>%
  ggplot()+
  geom_sf(aes(fill = CTYUA23NM ))+
  guides( fill = 'none')

ViceCounties <- st_read('../data/vice-counties-master/vice-counties-master/MLWM/County_Coastal_Terrestrial_region.shp')

ViceCounties %>%
  ggplot()+
  geom_sf(aes(fill = VCNAME  ))+
  guides( fill = 'none')

```

## Read in rainfall rasters

Login through CEDA

https://catalogue.ceda.ac.uk/uuid/18ddbb686be549bfadfecbe0c673f405/

Citable as:  Met Office; Hollis, D.; Carlisle, E.; Kendon, M.; Packman, S.; Doherty, A. (2024): HadUK-Grid Gridded Climate Observations on a 25km grid over the UK, v1.3.0.ceda (1836-2023). NERC EDS Centre for Environmental Data Analysis, 18 July 2024. doi:10.5285/18ddbb686be549bfadfecbe0c673f405. https://dx.doi.org/10.5285/18ddbb686be549bfadfecbe0c673f405


```{r}
### Loading Temps 
FilePathTo.nc<- "../data/dap.ceda.ac.uk/badc/ukmo-hadobs/data/insitu/MOHC/HadOBS/HadUK-Grid/v1.3.0.ceda/25km/tas/seas/v20240514/"
                 
Files<- list.files(FilePathTo.nc,full.names = TRUE)[-1] ## dropping 'index'

List_Rasters<- map(Files, rast)

List_Rasters[[2]] %>% plot

## ChangeNamesToTimes
List_Rasters2<- map(List_Rasters, function(RS){names(RS)<-  paste0('SEASON_', time(RS));return(RS)})

## Collapse the list
Collated_SeasonalTemps<- rast(List_Rasters2)
```

## Vice Counties

```{r echo=TRUE, include=FALSE}
VC_Seasonal_Temps<- exactextractr::exact_extract(Collated_SeasonalTemps,
                                                ViceCounties, fun = 'mean', weights ='area',
                                                append_cols = c( 'VCNUMBER','VCNAME' ))
```

```{r}

VC_Seasonal_Temps %>% 
  pivot_longer(cols = starts_with('mean'), 
               names_to = 'MidDate',
               values_to = 'MeanTemp',
               names_prefix = 'mean.SEASON_') %>%
  separate(MidDate     ,
           into = c( 'Year', 'Month', 'Mid' ),
           convert = TRUE, remove = FALSE)%>%
  mutate( Season = case_when(Month ==1 ~'Winter',
                             Month ==4 ~'Spring',
                             Month ==7 ~'Summer',
                             Month ==10 ~'Autumn')) -> VC_Seasonal_Temps_Long

write_csv(VC_Seasonal_Temps_Long, 
          '../data/VC_Seasonal_Temps_Long.csv')


## Visual checks

VC_Seasonal_Temps_Long %>%
  filter(VCNAME %in% ViceCounties$VCNAME[1:10])%>%
  ggplot(aes(col = VCNAME,  x= ymd(MidDate), y = MeanTemp ))+
  geom_line()+
  guides( col = 'none')


## mapping out a random year summer temps and comparing to raw data
ViceCounties %>%
  left_join( filter(VC_Seasonal_Temps_Long,
                    Year == 1992,
                    Season == 'Summer'),
             by = join_by(VCNUMBER, VCNAME)) %>%
  ggplot()+
  geom_sf(aes ( fill =MeanTemp  ))

ggplot()+
  geom_spatraster(data = List_Rasters[[109]], aes(fill = tas_3) )


```


## Unitary authoriries 


```{r echo=TRUE, include=FALSE}
UA_Seasonal_Temps<- exactextractr::exact_extract(Collated_SeasonalTemps,
                                                LegalCounties, fun = 'mean', weights ='area',
                                                append_cols = 'CTYUA23NM')
```

```{r}

UA_Seasonal_Temps %>% 
  pivot_longer(cols = starts_with('mean'), 
               names_to = 'MidDate',
               values_to = 'MeanTemp',
               names_prefix = 'mean.SEASON_') %>%
  separate(MidDate     ,
           into = c( 'Year', 'Month', 'Mid' ),
           convert = TRUE, remove = FALSE)%>%
  mutate( Season = case_when(Month ==1 ~'Winter',
                             Month ==4 ~'Spring',
                             Month ==7 ~'Summer',
                             Month ==10 ~'Autumn')) -> UA_Seasonal_Temps_Long

write_csv(UA_Seasonal_Temps_Long, 
          '../data/UA_Seasonal_Temps_Long.csv')


## Visual checks

UA_Seasonal_Temps_Long %>%
  filter(CTYUA23NM  %in% sample(LegalCounties$CTYUA23NM,10))%>%
  ggplot(aes(col = CTYUA23NM ,  x= ymd(MidDate), y = MeanTemp ))+
  geom_line()


## mapping out a random year summer temps and comparing to raw data
LegalCounties %>%
  left_join( filter(UA_Seasonal_Temps_Long,
                    Year == 1992,
                    Season == 'Summer'),
             by = 'CTYUA23NM') %>%
  ggplot()+
  geom_sf(aes ( fill =MeanTemp  ))

ggplot()+
  geom_spatraster(data = List_Rasters[[109]], aes(fill = tas_3) )

```

## Package Details

```{R}
sessionInfo()
```

