---
title: "Cleaning1_NHM"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r message = FALSE}
library(tidyverse)
library(readxl)
```

# Loading Data

## Measurements 

Using data from: Wilson et al https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13844

Sourced from: 
https://zenodo.org/records/5759759
and
https://zenodo.org/records/6066070

```{r}
WilsonSH3<- read_xlsx('../data/NHM/Wilson_et_al_MEE_data.xlsx', sheet = 3)
Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')
```

# Locations

Gordon L J Paterson; Sara Albuquerque; Vladimir Blagoderov; Stephen J Brooks et al. (2016). iCollections [Data set]. Natural History Museum. https://doi.org/10.5519/0038559

*DOI has been created for this download request: https://doi.org/10.5519/qd.tqng8ri6. Please ensure you reference this DOI when citing these data*

```{r eval = FALSE}
### Large files, so pre-wrangling data: 

Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')

NHM_data <- read_csv("C:/Users/pemb4504/Documents/7ba5ee8575924a906078531b9212aec880045725/resource.csv")

colnames(NHM_data)

NHM_data %>%
  select( "_id","basisOfRecord","catalogNumber",
          "class","collectionCode",
          "continent","coordinateUncertaintyInMeters", "country",                 
          "created","day","decimalLatitude","decimalLongitude" ,            
          "determinationFiledAs","determinationNames",  
          "family", "gbifID"    ,                   
          "gbifIssue", "genus",
          "higherClassification","month",      
          "order","project" ,                     
          "recordedBy" ,                 
          "scientificName",               
          "verbatimLatitude","verbatimLongitude","year" )-> NHM_data_firstcut  

ifelse(sapply(NHM_data_firstcut, function(x)all(is.na(x))) == TRUE, "Y","N")

NHM_data_firstcut %>%
  mutate( `specimen number` = parse_number(catalogNumber))  %>%
  filter(!is.na( decimalLatitude ))  %>%                            ## Excluding those without locations 
  select( catalogNumber,  `specimen number`,determinationNames ,
          decimalLatitude,decimalLongitude, year, month )  -> Just_Names_Loc_and_Numbers

write_csv(Just_Names_Loc_and_Numbers, '../data/Just_Names_Loc_and_Numbers.csv')
```

```{r warning=FALSE}
Sizes %>% 
  left_join(read_csv( '../data/Just_Names_Loc_and_Numbers.csv'), 
            by = join_by(`specimen number`)) -> Joined_sizes_and_Locs

```

# Processing

## First cut of filterring

Where wings differ by more than 2mm, assume either damage or AI errors and remove: 

```{r}
Joined_sizes_and_Locs%>%
  rename(wingspan_mm =`wing_span (mm)` ) %>%
  filter(!is.na( decimalLatitude ),  ## Excluding those without locations 
         wingspan_mm !=0, ## Exclude those without measurements  
         abs(`left_wing (mm)`-`right_wing (mm)`)<2, ## keep only those where the wing measurements are quite similiar  
         decimalLatitude <54, ## Exclude Northerly samples
         decimalLatitude> 49.5,  ### Exclude very southerly
         decimalLongitude > -5.5, ## Exclude Irish samples 
         year>1885 , ## Include only those where weather records are available
         wingspan_mm < 100) %>%   ## Exclude implausibly high measurements 
  group_by(`Taxon Name`)%>%
  add_count(name = 'Total_Obs') %>%
  filter( Total_Obs>50) %>%## Exclude species with very few observations
  ungroup -> NHM_filter1

```

## Aligning species names

```{r}
Names_table<- read_csv('../data/names_table.csv')

NHM_filter1 %>%
  rename(  `NHM name` = `Taxon Name` ) %>%
  left_join(Names_table, by = 'NHM name')%>%
  filter( !is.na(`Taxon name`))-> NHM_filter2  ## filter out those vagrants etc that are not in the name table because they are not going to be anaylsed

```

## Dealing with month windows by picking middle

```{r}
NHM_filter2 %>%
  mutate( month = case_when(month == '3 - 4'   ~ 3.5  ,
                            month == '5 - 9'   ~ 7  ,
                            month == '8 - 7'   ~ 7.5  ,
                            month == '6 - 8'   ~ 7  ,
                            month == '11 - 12' ~ 11.5  ,
                            month == '4 - 5'   ~ 4.5  ,
                            month == '8 - 9'   ~ 8.5  ,
                            month == '3 - 6'   ~ 4.5  ,
                            month == '6 - 9'   ~ 7.5   ,
                            month == '6 - 7'   ~ 6.5,
                            month == '5 - 6'   ~ 5.5  ,
                            month == '7 - 8'   ~ 7.5   , 
                            .default = as.numeric(month)))%>%
  filter( month >= 3) -> NHM_filter3       ########### FILTER OUT ALL from Jan and Feb


```

## Cutting down to focal columns:

```{r}
NHM_filter3 %>%
  mutate( LocationType = 'LatLong') %>%
  select( Individual = filename ,
          `Taxon name`,
          decimalLatitude, decimalLongitude, 
          LocationType ,
          Month =month, Year = year,
          Sex = gender,
          `left_wing (mm)`:`wing_shoulder (mm)` ) -> NHM_tidy

NHM_tidy
```

# Saving

```{r}
write_csv(NHM_tidy, '../data/NHM_measurements_clean.csv')
```

# Session info

```{r}
sessionInfo()
```
