---
title: "Cleaning4_OUMNH_Lats"
author: "Chris Terry & Elsa Heywood"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Adding the shapefiles to get midpoint data for each location and then combining that with temperature, brood, and name data.
# Load libraries
```{r message=FALSE}
library(sf)
library(tidyverse)
library(readxl)
```

# Read in shapefile data for vice counties and unitary authorities
```{r}
counties <- st_read("../data/vice-counties-master/vice-counties-master/12-mile/County_12mile_region.shp")
print(counties)

unitary <- st_read("../data/Counties_and_Unitary_Authorities_May_2023_UK_BGC_9070630924157476989/CTYUA_MAY_2023_UK_BGC.shp")
print(unitary)
```

# First extract midpoints, convert to LatLong for unitary authorities and clean
```{r}
# Get centroids of each unitary authority polygon
unitary_centroids <- st_centroid(unitary)

# Transform the centroids to WGS84 coordinate system (EPSG:4326, which is lat/lon)
unitary_centroids <- st_transform(unitary_centroids, crs = 4326)

# Extract coordinates into a data frame
centroid_coords <- st_coordinates(unitary_centroids)

# Combine coordinates with attribute data (like the name of each unitary authority)
unitary_midpoints <- unitary_centroids %>%
  st_drop_geometry() %>%
  bind_cols(as.data.frame(centroid_coords)) %>%
  rename(longitude = X, latitude = Y)

#clean columns
unitary_midpoints_clean <- unitary_midpoints %>%
  select(CTYUA23CD, CTYUA23NM, longitude, latitude)


#### this is all that is needed! 
unitary_midpoints_clean2 <- unitary %>%
  st_transform(crs = 4326)%>%
  st_centroid %>%
  mutate(LONG=st_coordinates(.)[,1],
         LAT =st_coordinates(.)[,2])%>%
  select(CTYUA23NM,LONG,LAT)%>%
  st_drop_geometry

unitary_midpoints_clean[1:5,]
unitary_midpoints_clean2[1:5,]

```

# View and save
```{r}
# View the result
#head(unitary_midpoints_clean)
#write_csv(unitary_midpoints_clean, "../data/unitary_authority_midpoints.csv")

```

# Same for vice counties
```{r}
##midpoints for counties
st_crs(counties)

# Get centroids
county_centroids <- st_centroid(counties)

# Transform to WGS84 (lat/lon)
county_centroids <- st_transform(county_centroids, crs = 4326)

# Extract coordinates
coords <- st_coordinates(county_centroids)

# Combine with attribute data
county_midpoints <- county_centroids %>%
  st_drop_geometry() %>%
  bind_cols(as.data.frame(coords)) %>%
  rename(longitude = X, latitude = Y)

#clean columns
county_midpoints_clean <- county_midpoints %>%
  select(VCNUMBER, VCNAME, longitude, latitude)

county_midpoints_clean2<- counties%>%
  st_centroid %>% 
  st_transform(crs = 4326)%>%
  mutate(LONG=st_coordinates(.)[,1],
         LAT =st_coordinates(.)[,2])%>%
  select(VCNAME,LONG,LAT)%>%
  st_drop_geometry
```

# View and save 
```{r}
# View result
#head(county_midpoints_clean)
#write_csv(county_midpoints_clean, "../data/county_midpoints.csv")

```

# Combine the unitary authority and vice county midpoint data
```{r}
# Add type to unitary authorities
unitary_midpoints_clean <- unitary_midpoints_clean %>%
  mutate(Location = CTYUA23NM,
         LocationType = "UA") %>%
  select(Location, LocationType, latitude, longitude)

# Add type to counties
county_midpoints_clean <- county_midpoints_clean %>%
  mutate(Location = VCNAME,
         LocationType = "VC") %>%
  select(Location, LocationType, latitude, longitude)
# Combine
combined_locations <- bind_rows(unitary_midpoints_clean, county_midpoints_clean)
head(combined_locations)
combined_locations %>% filter(LocationType == "VC") %>% head()


##round values to something sensible

combined_locations %>%
  mutate(latitude = round(latitude, digits = 6),
         longitude = round(longitude, digits = 6)) %>%
  write_csv("../data/combined_midpoints.csv")
```