---
title: "Cleaning3_Broods_and_Temps"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

This doc links each record with a temperature for that year's spring, summer and summer of the year before. 

The NHM data has lat-longs, the OUNHM data is at vice-county or unitary authority level. 


```{r}

library(tidyverse)
library(sf) # for polygons
library(tidyverse) # for general manipulation
library(terra) # for rasters
library(tidyterra) # for plotting rasters 
library(exactextractr)  ## for faster averaging of raster polygon intersections
```

# NHM Data

```{r}
NHM_records <- read_csv('../data/NHM_measurements_clean.csv')
```

Identify dates and locations that need to be crossreferenced:

```{r }
NHM_records %>%
  distinct(decimalLatitude, decimalLongitude, Year  ) %>%
mutate( 'SPRING'        =  paste0('SEASON_', Year, '-04-16'),
        'SUMMER'        =  paste0('SEASON_', Year, '-07-16'),
        'SUMMER_before' =  paste0('SEASON_', Year-1, '-07-16')) -> Long_Lats_Years
```

### Loading Temps 

Login through CEDA

https://catalogue.ceda.ac.uk/uuid/18ddbb686be549bfadfecbe0c673f405/

Citable as:  Met Office; Hollis, D.; Carlisle, E.; Kendon, M.; Packman, S.; Doherty, A. (2024): HadUK-Grid Gridded Climate Observations on a 25km grid over the UK, v1.3.0.ceda (1836-2023). NERC EDS Centre for Environmental Data Analysis, 18 July 2024. doi:10.5285/18ddbb686be549bfadfecbe0c673f405. https://dx.doi.org/10.5285/18ddbb686be549bfadfecbe0c673f405


```{r eval = FALSE}
FilePathTo.nc<- "../data/dap.ceda.ac.uk/badc/ukmo-hadobs/data/insitu/MOHC/HadOBS/HadUK-Grid/v1.3.0.ceda/25km/tas/seas/v20240514/"
Files<- list.files(FilePathTo.nc,full.names = TRUE)[-1] ## dropping 'index'
List_Rasters<- map(Files, rast)
## ChangeNamesToTimes
List_Rasters2<- map(List_Rasters, function(RS){names(RS)<-  paste0('SEASON_', time(RS));return(RS)})
## Collapse the list
Collated_SeasonalTemps<- rast(List_Rasters2)
```

## Matching Locations to Temperatures

```{r eval = FALSE}
UKGRID_COORDS<-sf_project(pts=  select(Long_Lats_Years,decimalLongitude,decimalLatitude ),
                          from ='epsg:4326', 
                          to ='+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +a=6377563.396 +rf=299.324961266495 +units=m +no_defs')

## Extracting all years out into a table of specimen x seasons
RecordTemps<- terra::extract(x = Collated_SeasonalTemps,
                             y=UKGRID_COORDS)
## Checking alignment is good
plot(Collated_SeasonalTemps[[1]] )
points( UKGRID_COORDS) 

## There will be some points that just miss a box. For them can setsearch_radius = 100, but not with multilayers
## Need to pull out RecordTemps

Spring<- rep(NA, nrow( RecordTemps))
Summer<- rep(NA, nrow( RecordTemps))
SummerBefore<- rep(NA, nrow( RecordTemps))

for( i in 1:nrow(RecordTemps)){
  ## dumb but not that slow way to get numbers out of matrix
  Spring[i] <- RecordTemps[i ,Long_Lats_Years$SPRING[i] ]
  Summer[i] <- RecordTemps[i ,Long_Lats_Years$SUMMER[i] ]
 SummerBefore[i] <- RecordTemps[i ,Long_Lats_Years$SUMMER_before[i] ]
}

Long_Lats_Years$SPRING_temp <- Spring
Long_Lats_Years$SUMMER_temp <- Summer
Long_Lats_Years$SUMMER_before_temp <- SummerBefore

```

### Identifying missed areas:
```{R eval = FALSE}
Long_Lats_Years %>%
  ggplot()+
  geom_point(aes( x = decimalLongitude, y = decimalLatitude,
                  col = is.na(Spring)))  ## Just weird coastal ones.
```

Fixing these using nearest neighbour temperatures

```{r eval = FALSE}
which(is.na(Long_Lats_Years$SPRING_temp)) -> WeirdCoastal_Numbs

for(i in WeirdCoastal_Numbs){   ## takes a little while to run (~1 minute)
  print(i)
  
  ## Spring
  SeasonLayer = Long_Lats_Years$SPRING[i] 
  yyy<- terra::extract(x = subset(Collated_SeasonalTemps,SeasonLayer) ,
                       y=vect(UKGRID_COORDS)[i],
                       search_radius=100000) ## up to 100km to nearest box
  Spring[i] <- yyy[1,2]
  
  ## Summer
  SeasonLayer = Long_Lats_Years$SUMMER[i] 
  xxx<- terra::extract(x = subset(Collated_SeasonalTemps,SeasonLayer) ,
                       y=vect(UKGRID_COORDS)[i],
                       search_radius=100000) ## up to 100km to nearest box
  Summer[i] <- xxx[1,2]
  
  ## Summer Before
    SeasonLayer = Long_Lats_Years$SUMMER_before[i] 
  zzz<- terra::extract(x = subset(Collated_SeasonalTemps,SeasonLayer) ,
                       y=vect(UKGRID_COORDS)[i],
                       search_radius=100000) ## up to 100km to nearest box
  SummerBefore[i] <- zzz[1,2]
  

}

## Replacing
Long_Lats_Years$SPRING_temp <- Spring
Long_Lats_Years$SUMMER_temp <- Summer
Long_Lats_Years$SUMMER_before_temp <- SummerBefore


### Checking all good now:
which(is.na(Long_Lats_Years$SPRING_temp))

```

### Saving

```{r eval = FALSE}
write_csv(Long_Lats_Years, 
          '../data/NHM_Long_Lats_Years.csv')

```


# OUNHM


## Formatting Climate Data

### Read in County shape files

Unitary Authorities:

https://www.data.gov.uk/dataset/85228aec-fe0e-49bf-9455-df000d61e731/counties-and-unitary-authorities-may-2023-boundaries-uk-bgc

Vice Counties 

From https://www.brc.ac.uk/article/british-vice-counties / https://github.com/BiologicalRecordsCentre/vice-counties/tree/master

```{r message=FALSE}
LegalCounties<- st_read('../data/Counties_and_Unitary_Authorities_May_2023_UK_BGC_9070630924157476989/CTYUA_MAY_2023_UK_BGC.shp')

LegalCounties %>%
  ggplot()+
  geom_sf(aes(fill = CTYUA23NM ))+
  guides( fill = 'none')

ViceCounties <- st_read('../data/vice-counties-master/vice-counties-master/MLWM/County_Coastal_Terrestrial_region.shp')

ViceCounties %>%
  ggplot()+
  geom_sf(aes(fill = VCNAME  ))+
  guides( fill = 'none')
```

## Vice Counties

```{r echo=TRUE, include=FALSE}
VC_Seasonal_Temps<- exactextractr::exact_extract(Collated_SeasonalTemps,
                                                ViceCounties, fun = 'mean', weights ='area',
                                                append_cols = c( 'VCNUMBER','VCNAME' ))
```

```{r}
VC_Seasonal_Temps %>% 
  pivot_longer(cols = starts_with('mean'), 
               names_to = 'MidDate',
               values_to = 'MeanTemp',
               names_prefix = 'mean.SEASON_') %>%
  separate(MidDate     ,
           into = c( 'Year', 'Month', 'Mid' ),
           convert = TRUE, remove = FALSE)%>%
  mutate( Season = case_when(Month ==1 ~'Winter',
                             Month ==4 ~'Spring',
                             Month ==7 ~'Summer',
                             Month ==10 ~'Autumn')) -> VC_Seasonal_Temps_Long

write_csv(VC_Seasonal_Temps_Long, 
          '../data/VC_Seasonal_Temps_Long.csv')

## Visual checks

VC_Seasonal_Temps_Long %>%
  filter(VCNAME %in% ViceCounties$VCNAME[1:10])%>%
  ggplot(aes(col = VCNAME,  x= ymd(MidDate), y = MeanTemp ))+
  geom_line()+
  guides( col = 'none')


## mapping out a random year summer temps and comparing to raw data
ViceCounties %>%
  left_join( filter(VC_Seasonal_Temps_Long,
                    Year == 1992,
                    Season == 'Summer'),
             by = join_by(VCNUMBER, VCNAME)) %>%
  ggplot()+
  geom_sf(aes ( fill =MeanTemp  ))

ggplot()+
  geom_spatraster(data = List_Rasters[[109]], aes(fill = tas_3) )


```

## Unitary authoriries 

```{r echo=TRUE, include=FALSE}
UA_Seasonal_Temps<- exactextractr::exact_extract(Collated_SeasonalTemps,
                                                LegalCounties, fun = 'mean', weights ='area',
                                                append_cols = 'CTYUA23NM')
```

```{r}

UA_Seasonal_Temps %>% 
  pivot_longer(cols = starts_with('mean'), 
               names_to = 'MidDate',
               values_to = 'MeanTemp',
               names_prefix = 'mean.SEASON_') %>%
  separate(MidDate     ,
           into = c( 'Year', 'Month', 'Mid' ),
           convert = TRUE, remove = FALSE)%>%
  mutate( Season = case_when(Month ==1 ~'Winter',
                             Month ==4 ~'Spring',
                             Month ==7 ~'Summer',
                             Month ==10 ~'Autumn')) -> UA_Seasonal_Temps_Long

write_csv(UA_Seasonal_Temps_Long, 
          '../data/UA_Seasonal_Temps_Long.csv')

## Visual checks

UA_Seasonal_Temps_Long %>%
  filter(CTYUA23NM  %in% sample(LegalCounties$CTYUA23NM,10))%>%
  ggplot(aes(col = CTYUA23NM ,  x= ymd(MidDate), y = MeanTemp ))+
  geom_line()


## mapping out a random year summer temps and comparing to raw data
LegalCounties %>%
  left_join( filter(UA_Seasonal_Temps_Long,
                    Year == 1992,
                    Season == 'Summer'),
             by = 'CTYUA23NM') %>%
  ggplot()+
  geom_sf(aes ( fill =MeanTemp  ))

ggplot()+
  geom_spatraster(data = List_Rasters[[109]], aes(fill = tas_3) )

```

# Oxford Measurements

```{r}
OUNHM_data <- read_csv('../data/OUNHM_measurements_clean.csv')
UA_temps <- read_csv( '../data/UA_Seasonal_Temps_Long.csv') 
VC_temps <- read_csv('../data/VC_Seasonal_Temps_Long.csv')
```

```{r}

UA_temps %>%
  select(Location =  CTYUA23NM  , Year, Season, MeanTemp ) %>%
    mutate(LocationType= 'UA') -> UA_tidy


VC_temps %>%
  select(Location =  VCNAME         , Year, Season, MeanTemp ) %>%
    mutate(LocationType= 'VC') -> VC_tidy

bind_rows(UA_tidy, VC_tidy) -> Both_Temps


Both_Temps %>%
  filter( Season %in% c( 'Spring', 'Summer')) %>%
  group_by(Location, LocationType, Year) %>%
  pivot_wider(id_cols = c('Location', 'LocationType', 'Year'),
              names_from = 'Season', values_from = 'MeanTemp') -> SpringSummerTemps

## Calculating YearBefore

SpringSummerTemps %>%
  select(Summer_before = Summer)%>%
  mutate(Year = Year+1) -> Summer_Before


SpringSummerTemps %>% 
  left_join(Summer_Before,
            by = c('Location', 'LocationType', 'Year')) %>%
  filter( !is.na(Summer_before)) -> ThreeTemps_perLocYear


OUNHM_data %>%
  left_join(ThreeTemps_perLocYear,
            by = c('Location', 'LocationType', 'Year'))

```


























