---
title: "Monthly Data"
output: html_document
author: "Heywood et al."
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document links each butterfly record with a temperature in months over the preceding year or so.  

The NHM data has lat-longs, the OUMNH data is at vice-county or unitary authority level, so requires some different handling.


```{r}
library(tidyverse)
library(sf) # for polygons
library(tidyverse) # for general manipulation
library(terra) # for rasters
library(tidyterra) # for plotting rasters 
library(exactextractr)  ## for faster averaging of raster polygon intersections
```

# NHM Data

```{r}
NHM_records <- read_csv('../data/NHM_measurements_clean.csv')
```

Identify dates and locations that need to be cross-referenced:

```{r }
NHM_records %>%
  distinct(decimalLatitude, decimalLongitude, Year  ) -> Long_Lats_Years  # unique locations and year
```

### Loading Temps 

Login through CEDA

```{r}
#'/badc/ukmo-hadobs/data/insitu/MOHC/HadOBS/HadUK-Grid/v1.3.1.ceda/25km/tas/mon/v20250415'
# wget -O -q https://dap.ceda.ac.uk/badc/ARCHIVE_INFO/ACCESS_TEST/RESTRICTED/TOKEN_CHECK --header "Authorization: Bearer TOKEN"

## token from 
#wget -e robots=off --mirror --no-parent -r https://dap.ceda.ac.uk/badc/ukmo-hadobs/data/insitu/MOHC/HadOBS/HadUK-Grid/v1.3.1.ceda/25km/tas/mon/ --header "Authorization: Bearer TOKEN"
```

https://catalogue.ceda.ac.uk/uuid/18ddbb686be549bfadfecbe0c673f405/

Citable as:  Met Office; Hollis, D.; Carlisle, E.; Kendon, M.; Packman, S.; Doherty, A. (2024): HadUK-Grid Gridded Climate Observations on a 25km grid over the UK, v1.3.0.ceda (1836-2023). NERC EDS Centre for Environmental Data Analysis, 18 July 2024. doi:10.5285/18ddbb686be549bfadfecbe0c673f405. https://dx.doi.org/10.5285/18ddbb686be549bfadfecbe0c673f405


```{r eval = FALSE}
FilePathTo.nc<- "../data/dap.ceda.ac.uk/badc/ukmo-hadobs/data/insitu/MOHC/HadOBS/HadUK-Grid/v1.3.1.ceda/25km/tas/mon/v20250415/"
Files<- list.files(FilePathTo.nc,full.names = TRUE)[-1] ## dropping 'index'
List_Rasters<- map(Files, rast)
## ChangeNamesToTimes
List_Rasters2<- map(List_Rasters, function(RS){names(RS)<-  paste0('MONTH_', time(RS));return(RS)})
## Collapse the list
Collated_MonthlyTemps<- rast(List_Rasters2)
```


## Matching Locations to Temperatures

```{r eval = FALSE}

## just select relevant area
UKGRID_COORDS<-sf_project(pts=  select(Long_Lats_Years,decimalLongitude,decimalLatitude ),
                          from ='epsg:4326', 
                          to ='+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +a=6377563.396 +rf=299.324961266495 +units=m +no_defs')

## Extracting all years out into a table of specimen x seasons
RecordTemps<- terra::extract(x = Collated_MonthlyTemps,
                             y=UKGRID_COORDS)


## Checking alignment is good
# plot(Collated_SeasonalTemps[[1]] )
# points( UKGRID_COORDS) 

August      <- rep(NA, nrow( RecordTemps))
July        <- rep(NA, nrow( RecordTemps))
June        <- rep(NA, nrow( RecordTemps))
May         <- rep(NA, nrow( RecordTemps))
April       <- rep(NA, nrow( RecordTemps))
March       <- rep(NA, nrow( RecordTemps))
SeptBefore  <- rep(NA, nrow( RecordTemps))
AugustBefore<- rep(NA, nrow( RecordTemps))
JulyBefore  <- rep(NA, nrow( RecordTemps))
JuneBefore  <- rep(NA, nrow( RecordTemps))

## Build Names of columns to pluck:

Long_Lats_Years$August      <-paste0('MONTH_', Long_Lats_Years$Year  , '-08-16' ) 
Long_Lats_Years$July        <-paste0('MONTH_', Long_Lats_Years$Year  , '-07-16' )
Long_Lats_Years$June        <-paste0('MONTH_', Long_Lats_Years$Year  , '-06-16' )
Long_Lats_Years$May         <-paste0('MONTH_', Long_Lats_Years$Year  , '-05-16' )
Long_Lats_Years$April       <-paste0('MONTH_', Long_Lats_Years$Year  , '-04-16' )
Long_Lats_Years$March       <-paste0('MONTH_', Long_Lats_Years$Year  , '-03-16' )
Long_Lats_Years$SeptBefore  <-paste0('MONTH_', Long_Lats_Years$Year-1, '-09-16' )
Long_Lats_Years$AugustBefore<-paste0('MONTH_', Long_Lats_Years$Year-1, '-08-16' )
Long_Lats_Years$JulyBefore  <-paste0('MONTH_', Long_Lats_Years$Year-1, '-07-16' )
Long_Lats_Years$JuneBefore  <-paste0('MONTH_', Long_Lats_Years$Year-1, '-06-16' )


for( i in 1:nrow(RecordTemps)){
  ## dumb but not that slow way to get numbers out of matrix
  August      [i]<-    RecordTemps[i ,Long_Lats_Years$August      [i]]
  July        [i]<-    RecordTemps[i ,Long_Lats_Years$July        [i]]
  June        [i]<-    RecordTemps[i ,Long_Lats_Years$June        [i]]
  May         [i]<-    RecordTemps[i ,Long_Lats_Years$May         [i]]
  April       [i]<-    RecordTemps[i ,Long_Lats_Years$April       [i]]
  March       [i]<-    RecordTemps[i ,Long_Lats_Years$March       [i]]
  SeptBefore  [i]<-    RecordTemps[i ,Long_Lats_Years$SeptBefore  [i]]
  AugustBefore[i]<-    RecordTemps[i ,Long_Lats_Years$AugustBefore[i]]
  JulyBefore  [i]<-    RecordTemps[i ,Long_Lats_Years$JulyBefore  [i]]
  JuneBefore  [i]<-    RecordTemps[i ,Long_Lats_Years$JuneBefore  [i]]
}

Long_Lats_Years$T_August      <- August     
Long_Lats_Years$T_July        <- July       
Long_Lats_Years$T_June        <- June       
Long_Lats_Years$T_May         <- May        
Long_Lats_Years$T_April       <- April      
Long_Lats_Years$T_March       <- March      
Long_Lats_Years$T_SeptBefore  <- SeptBefore
Long_Lats_Years$T_AugustBefore<- AugustBefore
Long_Lats_Years$T_JulyBefore  <- JulyBefore  
Long_Lats_Years$T_JuneBefore  <- JuneBefore  

```


### Identifying missed areas:
```{R eval = FALSE}
Long_Lats_Years %>%
  ggplot()+
  geom_point(aes( x = decimalLongitude, y = decimalLatitude,
                  col = is.na(T_July)))  ## Just weird coastal ones.
```

Fixing these using nearest neighbour temperatures

```{r eval = FALSE}
which(is.na(Long_Lats_Years$T_May)) -> WeirdCoastal_Numbs
Months<- c('August','July','June','May' ,'April','March',
           'SeptBefore','AugustBefore','JulyBefore','JuneBefore'  )

Long_Lats_Years<- as.data.frame(Long_Lats_Years) ## to allow slotting of values

for(i in WeirdCoastal_Numbs){   ## takes a little while to run (~1 minute)
  print(i)
  
  XX<- rep(NA, length(Months))
  
  for(m in 1:length(Months)){
    SeasonLayer = Long_Lats_Years[[i,Months[m]] ]
    
    x<- terra::extract(x =subset(Collated_MonthlyTemps,
                                 SeasonLayer) ,
                       y=vect(UKGRID_COORDS)[i],
                       search_radius=50000) ## up to 50km to nearest box
    XX[m]<- x[1,2]
  }
  Long_Lats_Years[i, 14:23] <- XX
}

Long_Lats_Years<- as_tibble(Long_Lats_Years) ## to allow slotting of values

### Checking all good now:
which(is.na(Long_Lats_Years$T_April))
which(is.na(Long_Lats_Years$T_JulyBefore))


```

### Saving

```{r eval = FALSE}
write_csv(Long_Lats_Years, 
          '../data/NHM_Long_Lats_Years.csv')
```


```{r}
Long_Lats_Years <- read_csv(  '../data/NHM_Long_Lats_Years.csv')

NHM_records %>%
  left_join(select( Long_Lats_Years,
                    decimalLatitude,
                    decimalLongitude,
                    Year , 
                    starts_with('T_')),
            by = c('decimalLatitude', 'decimalLongitude', 'Year')) %>%
  write_csv('../data/NHM_w_temps_Monthly.csv')

```


# OUMNH records


## Formatting Climate Data

### Read in County shape files

Unitary Authorities:

https://www.data.gov.uk/dataset/85228aec-fe0e-49bf-9455-df000d61e731/counties-and-unitary-authorities-may-2023-boundaries-uk-bgc

Vice Counties 

From https://www.brc.ac.uk/article/british-vice-counties / https://github.com/BiologicalRecordsCentre/vice-counties/tree/master

```{r message=FALSE}
LegalCounties<- st_read('../data/Counties_and_Unitary_Authorities_May_2023_UK_BGC_9070630924157476989/CTYUA_MAY_2023_UK_BGC.shp')

LegalCounties %>%
  ggplot()+
  geom_sf(aes(fill = CTYUA23NM ))+
  guides( fill = 'none')

ViceCounties <- st_read('../data/vice-counties-master/vice-counties-master/MLWM/County_Coastal_Terrestrial_region.shp')

ViceCounties %>%
  ggplot()+
  geom_sf(aes(fill = VCNAME  ))+
  guides( fill = 'none')
```

### Vice Counties

```{r echo=TRUE, include=FALSE, eval = FALSE}
VC_Seasonal_Temps<- exactextractr::exact_extract(Collated_MonthlyTemps,
                                                 ViceCounties, fun = 'mean', weights ='area',
                                                 append_cols = c( 'VCNUMBER','VCNAME' ))

write_csv(VC_Seasonal_Temps, '../data/VC_Monthly_Temps.csv' )
```

```{r}
VC_Seasonal_Temps<- read_csv('../data/VC_Monthly_Temps.csv' )

VC_Seasonal_Temps %>% 
  pivot_longer(cols = starts_with('mean'), 
               names_to = 'MidDate',
               values_to = 'MeanTemp',
               names_prefix = 'mean.MONTH_') %>%
  separate(MidDate     ,
           into = c('Year', 'Month', 'MidDay' ),
           convert = TRUE, remove = FALSE)%>%
  select( VCNUMBER,VCNAME, Year, Month, MeanTemp ) %>%
  mutate( MeanTemp = round( MeanTemp, 6)) -> VC_Monthly_Temps_Long

write_csv(VC_Monthly_Temps_Long, 
          '../data/VC_Monthly_Temps_Long.csv')

## Visual checks

## mapping out a random year summer temps and comparing to raw data
ViceCounties %>%
  left_join( filter(VC_Monthly_Temps_Long,
                    Year == 1992,
                    Month ==6),
             by = join_by(VCNUMBER, VCNAME)) %>%
  ggplot()+
  geom_sf(aes ( fill =MeanTemp  ))

#ggplot()+ geom_spatraster(data = List_Rasters[[109]], aes(fill = tas_3) )

```

### Unitary authoriries 

```{r echo=TRUE, eval=FALSE}
UA_Monthly_Temps<- exactextractr::exact_extract(Collated_MonthlyTemps,
                                                LegalCounties, fun = 'mean', weights ='area',
                                                append_cols = 'CTYUA23NM')

UA_Monthly_Temps %>% 
  pivot_longer(cols = starts_with('mean'), 
               names_to = 'MidDate',
               values_to = 'MeanTemp',
               names_prefix = 'mean.MONTH_') %>%
  separate(MidDate     ,
           into = c('Year', 'Month', 'MidDay' ),
           convert = TRUE, remove = FALSE) %>%
  mutate( MeanTemp = round( MeanTemp, 6)) -> UA_Monthly_Temps_Long

write_csv(UA_Monthly_Temps_Long, 
          '../data/UA_Monthly_Temps_Long.csv')
```

```{r}
UA_Monthly_Temps_Long <- read_csv( '../data/UA_Monthly_Temps_Long.csv')

## Visual checks

UA_Monthly_Temps_Long %>%
  filter(CTYUA23NM  %in% sample(LegalCounties$CTYUA23NM,10))%>%
  ggplot(aes(col = CTYUA23NM ,  x= ymd(MidDate), y = MeanTemp ))+
  geom_line()


## mapping out a random year summer temps and comparing to raw data
LegalCounties %>%
  left_join( filter(UA_Monthly_Temps_Long,
                    Year == 1992,
                    Month==6),
             by = 'CTYUA23NM') %>%
  ggplot()+
  geom_sf(aes ( fill =MeanTemp  ))

#ggplot()+ geom_spatraster(data = List_Rasters[[109]], aes(fill = tas_3) )

```

## Oxford Measurements

```{r}
OUMNH_data <- read_csv('../data/OUMNH_measurements_clean.csv')
UA_temps <- read_csv( '../data/UA_Monthly_Temps_Long.csv') 
VC_temps <- read_csv('../data/VC_Monthly_Temps_Long.csv')
```

```{r}

UA_temps %>%
  select(Location =  CTYUA23NM  , Year, Month, MeanTemp ) %>%
  mutate(LocationType= 'UA') -> UA_tidy


VC_temps %>%
  select(Location =  VCNAME         , Year, Month, MeanTemp ) %>%
  mutate(LocationType= 'VC') -> VC_tidy

bind_rows(UA_tidy, VC_tidy) -> Both_Temps


Both_Temps %>%
  filter( Month>2) %>%
  group_by(Location, LocationType, Year) %>%
  pivot_wider(id_cols = c('Location', 'LocationType', 'Year'),
              names_from = 'Month', values_from = 'MeanTemp') -> MonthlyTemps

## Calculating YearBefore

MonthlyTemps %>%
  select(SeptBefore = `9`,
         AugustBefore = `8`,
         JulyBefore   = `7`,
         JuneBefore   = `6`)%>%
  mutate(Year = Year+1) -> BeforeMonths


MonthlyTemps %>% 
  left_join(BeforeMonths,
            by = c('Location', 'LocationType', 'Year')) %>%
  filter( !is.na(SeptBefore))%>%
  select( T_August   =   `8` ,
          T_July    = `7`   ,
          T_June    =   `6`,
          T_May  =     `5`,
          T_April    =  `4`,
          T_March    =   `3`  ,
          T_SeptBefore = SeptBefore ,
          T_AugustBefore = AugustBefore,
          T_JulyBefore   = JulyBefore  ,
          T_JuneBefore   = JuneBefore  )-> Temps_perLocYear

OUMNH_data %>%
  left_join(Temps_perLocYear,
            by = c('Location', 'LocationType', 'Year'))-> OUMNH_data_temps

```

### Saving

```{r}

write_csv(OUMNH_data_temps, '../data/OUMNH_data_temps_Monthly.csv')

```

