---
title: "Cleaning1_NHM"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r message = FALSE}
library(tidyverse)
library(readxl)
```

# Loading Data

## Measurements 

Using data from: Wilson et al https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13844

Sourced from: 
https://zenodo.org/records/5759759
and
https://zenodo.org/records/6066070

```{r}

BroodTable <- read_csv('../data/brood_table.csv')

Names_table<- read_csv('../data/names_table.csv')

BroodTable %>%
  left_join(Names_table, by = 'NHM name')%>% 
  mutate(FirstMonth = as.numeric(FirstMonthBrood_focal ),
         LastMonth =  as.numeric(LastMonthBrood_focal )) %>%
  filter(Include == 'TRUE' )%>%
  select(`Taxon name` ,  BroodFocalUse, FirstMonth,LastMonth, Univoltine )-> Brood_table2

```


# Locations

Gordon L J Paterson; Sara Albuquerque; Vladimir Blagoderov; Stephen J Brooks et al. (2016). iCollections [Data set]. Natural History Museum. https://doi.org/10.5519/0038559

*DOI has been created for this download request: https://doi.org/10.5519/qd.tqng8ri6. Please ensure you reference this DOI when citing these data*

```{r}

WilsonSH3<- read_xlsx('../data/NHM/Wilson_et_al_MEE_data.xlsx', sheet = 3)

Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')

Sizes %>% 
  left_join(read_csv( '../data/Just_Names_Loc_and_Numbers.csv'), 
            by = join_by(`specimen number`)) -> Joined_sizes_and_Locs
```

# Adding Brood Tables to NHM
## Now add temperature to use column from   Brood_table_long
```{r}
## Which brood is each record
read_csv('../data/NHM_w_temps.csv')%>%
  left_join(select(Brood_table2,
                   `Taxon name`, FirstMonth,LastMonth, Univoltine,
                   BroodFocalUse ), 
            by = 'Taxon name') %>%
  mutate(InFocalBrood = ifelse(Univoltine, TRUE, 
                               ifelse( between(Month,
                                               FirstMonth,
                                               LastMonth),
                                       TRUE, FALSE))) %>%
  filter( InFocalBrood) %>% 
  filter(!Implausible ) %>%
  select(filename = Individual                , 
         Sp = `Taxon name`,
         Latitude= decimalLatitude ,
         Longitude= decimalLongitude  ,
         Month, Year ,Sex  , 
         Right_wing=`right_wing (mm)` ,
         BroodFocalUse  ,
         SPRING_temp,
         SUMMER_temp,
         SUMMER_before_temp  ) %>%
  mutate( Temp = case_when( BroodFocalUse == 'Spring' ~ SPRING_temp,
                            BroodFocalUse == 'Summer' ~ SUMMER_temp,
                            BroodFocalUse == 'Summer_before' ~ SUMMER_before_temp,
                            TRUE ~ NA_real_),
          Temp = round(Temp, digits = 3),
          Sex = ifelse( Sex =='N/A', 'undetermined', Sex)   ) %>%
  mutate(YearRescale = (Year-1800)/100) %>%
  select(-ends_with('_temp')) -> NHM_for_models




WilsonSH3 %>%
  select(ID, 
         Wilson_Sex = Sex   ,
         Wilson_Species = Species,
         Wilson_LateTemp = LateTemp,
         Wilson_Year =Year,
         Wilson_LnAv = `ln(av)`,
         Wilson_Length =average_forewing_length  ) ->Wilsonfocal

### Checking duplicates
left_join(NHM_for_models,
          select( Sizes, `specimen number`,  filename ),  by = 'filename' )[5457,] 

## NB some duplicates
Wilsonfocal %>%
  filter( ID ==1178561)

Wilsonfocal_D <-distinct(Wilsonfocal )


left_join(NHM_for_models,
          select( Sizes, `specimen number`,  filename ),  by = 'filename' ) %>%
  left_join(Wilsonfocal_D, by = c(`specimen number`  = 'ID')) -> CompareStudies

```

```{r}
CompareStudies %>%
  count(is.na(Wilson_Species))  ## something like 14k /90k  overlap

CompareStudies %>%
  filter( !is.na(Wilson_Species )) %>%
  ggplot( aes( Temp , Wilson_LateTemp ))+
  geom_point()+
  facet_wrap(~Sp)+
  coord_fixed()

## somewhat positive correlation in temperatures used

## Wilson has more variation due to month specificity
## We have more variation due to different regions


## just to check essentially same data 
CompareStudies %>%
  ggplot( aes( x = log(Right_wing), y = Wilson_LnAv, col = Sp))+
  geom_point()



```

## Does including 'year' remove Wilson's Temperature effects?

```{r}

Fits <- distinct(Wilsonfocal_D, Wilson_Sex, Wilson_Species)


for( i in 1: nrow( Fits)){
  
  CompareStudies %>% 
    filter(Wilson_Species == Fits$Wilson_Species[i], 
           Wilson_Sex     == Fits$Wilson_Sex[i] ) -> focal_data
  ### Wilson Data fits
  temp_model      <- summary(lm(Wilson_LnAv  ~ Wilson_LateTemp, data = focal_data))
  temp_year_model <- summary(lm(Wilson_LnAv  ~ Wilson_LateTemp + Wilson_Year, data = focal_data))
  
  Fits$Temp_p_solo[i] <- temp_model$coefficients[2,4] 
  Fits$Temp_p_wYear[i] <- temp_year_model$coefficients[2,4] 
  
  Fits$Temp_coef_solo[i] <- temp_model$coefficients[2,1] 
  Fits$Temp_coef_wYear[i] <- temp_year_model$coefficients[2,1] 
  
  Fits$WilsonR2_temp[i] = temp_model$r.squared

  
 ## Comparison with same data using Oxford temps
  temp_model      <- summary(lm(Wilson_LnAv  ~  Temp , data = focal_data))
  temp_year_model <- summary(lm(Wilson_LnAv ~  Temp  + Year , data = focal_data))
  
  Fits$OX_Temp_p_solo[i] <- temp_model$coefficients[2,4] 
  Fits$OX_Temp_p_wYear[i] <- temp_year_model$coefficients[2,4] 
  
  Fits$OX_Temp_coef_solo[i] <- temp_model$coefficients[2,1] 
  Fits$OX_Temp_coef_wYear[i] <- temp_year_model$coefficients[2,1] 
  
  
  Fits$OxR2_temp[i] = temp_model$r.squared
  
  
  Fits$N[i] <- nrow( focal_data)
}






Fits %>%
  ggplot( aes( x = Temp_p_solo,y= Temp_p_wYear))+
  geom_point()

Fits %>%
  ggplot( aes( x = Temp_coef_solo,y= Temp_coef_wYear))+
  geom_point()

Fits %>%
  ggplot( aes( x = WilsonR2_temp ,y= OxR2_temp , col = log10(N)))+
  geom_point()

Fits %>%
  arrange( desc(OxR2_temp))

## Strongest signal of oxford temperature is with male H semele

 CompareStudies %>% 
    filter(Wilson_Species ==  'H semele'    , 
           Wilson_Sex     == 'M') %>%
   ggplot(aes( x = Temp, y = Right_wing, col = Latitude))+
   geom_point()
 
 ### but very correlated with latitude:
  CompareStudies %>% 
    filter(Wilson_Species ==  'H semele'    , 
           Wilson_Sex     == 'M') %>%
   ggplot(aes( x = Latitude , y = Right_wing, col = Temp))+
   geom_point()

  
  
  ## not seen in Wilson data
  
   CompareStudies %>% 
    filter(Wilson_Species ==  'H semele'    , 
           Wilson_Sex     == 'M') %>%
   ggplot(aes( x = Wilson_LateTemp, y = Right_wing, col = Latitude))+
   geom_point()
  
## But they do report in their table, but opposing trends depending on stage
#                                Change in size with temperature in:
#Species	     Sex  Late larval stage (%)	Early larval stage (%) Pupal stage (%)
#H. semele 	Male     	0.40	                  -0.39                 	0.85

   
   
   
  
  
  


Fits %>%
  ggplot( aes( x = OX_Temp_coef_solo,y= Temp_coef_solo))+
  geom_point()


Fits %>%
  ggplot( aes( x = OX_Temp_coef_wYear,y= Temp_coef_wYear))+
  geom_point()

```
