---
title: "Matching Records to Temperatures"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r}

library(tidyverse)
library(sf) # for polygons
library(tidyverse) # for general manipulation
library(terra) # for rasters
library(tidyterra) # for plotting rasters 
library(exactextractr)  ## for faster avaerging of raster polygon intersections

```

# Purpose

This file matches the temperature record data to the NHM and OUNHM butterfly records by location

# Targets


## NHM 

```{r}
NHM_records <- read_csv('../data/Just_Names_Loc_and_Numbers.csv')
```

### Cleaning up years

A fair few of the NHM records are listed as having dates into the future. Seems optimistic.

Checking a random subset of these confirms mistake:

https://data.nhm.ac.uk/dataset/56e711e6-c847-4f99-915a-6894bb5c5dea/resource/05ff2255-c38a-40c9-b657-4ccb55ab2feb/record/6989349

https://data.nhm.ac.uk/dataset/56e711e6-c847-4f99-915a-6894bb5c5dea/resource/05ff2255-c38a-40c9-b657-4ccb55ab2feb/record/6769801

https://data.nhm.ac.uk/dataset/56e711e6-c847-4f99-915a-6894bb5c5dea/resource/05ff2255-c38a-40c9-b657-4ccb55ab2feb/record/6930208

Where to draw cutoff??

This example from 2024: 
https://data.nhm.ac.uk/dataset/56e711e6-c847-4f99-915a-6894bb5c5dea/resource/05ff2255-c38a-40c9-b657-4ccb55ab2feb/record/6781405

is clearly meant to be 1924 (just '24 n the label, but in a very old pin)


https://data.nhm.ac.uk/dataset/56e711e6-c847-4f99-915a-6894bb5c5dea/resource/05ff2255-c38a-40c9-b657-4ccb55ab2feb/record/7055394

Problem goes back as far as 2011

https://data.nhm.ac.uk/dataset/56e711e6-c847-4f99-915a-6894bb5c5dea/resource/05ff2255-c38a-40c9-b657-4ccb55ab2feb/record/7012818


----

https://data.nhm.ac.uk/dataset/icollections

says the database extends *from 1819 until the 1980s*. https://bdj.pensoft.net/article/9559/ includes a histogram of dates that seems plausible https://doi.org/10.3897/BDJ.4.e9559.figure12 but doesn't match:

How dd Wilson et al deal with this? Seems that they didn't run their AI over the future measurements? OR has some other method
```{r}
#  NHM_records%>% 
#   filter(year>2016) %>%
#   mutate(catalogNumber= parse_number(catalogNumber)) %>%
#   pluck('catalogNumber'  ) -> dodgeyCatNums
# 
#   WilsonSH3<- read_xlsx('../data/NHM/Wilson_et_al_MEE_data.xlsx', sheet = 3)
#   Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')
# 
# Sizes %>%
#   filter(`specimen number` %in% dodgeyCatNums)     #### Seems like the filter them out?
```


```{r}
NHM_records$year %>% hist ; abline(v=2025)
```

```{R}
FilteredSizeAndLoc<-read_csv('../data/FilteredSizeAndLoc.csv') 
FilteredSizeAndLoc$year %>% hist ; abline(v=2025)
```

Locations are given by decimal lats and longs not names


### Aligning species names

```{r}
BroodTable<-readxl::read_excel( '../data/butterfly brood table.xlsx')
NHM_SpeciesNames <- unique(FilteredSizeAndLoc$`Taxon Name`)
all(NHM_SpeciesNames %in% BroodTable$`Taxonomic species`) ## check all species in table
```


### Splitting broods and defining seasons to use

```{r}
BroodTable %>%
  filter( Include == 'TRUE', Univoltine) %>%
  select(`Taxonomic species`,
         TempDEVELOP=Brood1_TempDEVELOP,
         TempLAY= Brood1_TempLAY) %>%
  mutate( Brood = 'Univoltine') -> UNIVOLT_sp

BroodTable %>%
  filter( Include == 'TRUE', !Univoltine) %>%
  select(`Taxonomic species`,Brood1_TempDEVELOP:Brood2_TempLAY)%>%
  pivot_longer(cols =  -`Taxonomic species`,
               names_to = 'Brood',
               values_to = 'Season') %>%
  separate(Brood , into = c('Brood', 'Batch')) %>%
  pivot_wider(id_cols =  c(`Taxonomic species`,Brood   ),
              names_from = 'Batch'  , values_from =  'Season'  )    -> BIVOLT_sp     

AllBroodTable_Seasons <- bind_rows( UNIVOLT_sp, BIVOLT_sp) %>%
  rename( `Taxon Name` = `Taxonomic species`) 
```

```{r }
## Filterring and simplfying

FilteredSizeAndLoc %>%
  filter(!is.na(year)) %>%
  filter(year > 1885) %>%
  select(`Taxon Name` ,month,  year, decimalLatitude, decimalLongitude ,
         `specimen number`,gender, wingspan_mm ) -> NHM_filterred

### Dealing with weird months

NHM_filterred %>%
  mutate( month = case_when(month == '3 - 4'   ~ 3.5  ,
                            month == '5 - 9'   ~ 7  ,
                            month == '8 - 7'   ~ 7.5  ,
                            month == '6 - 8'   ~ 7  ,
                            month == '11 - 12' ~ 11.5  ,
                            month == '4 - 5'   ~ 4.5  ,
                            month == '8 - 9'   ~ 8.5  ,
                            month == '3 - 6'   ~ 4.5  ,
                            month == '6 - 9'   ~ 7.5   ,
                            month == '6 - 7'   ~ 6.5,
                            month == '5 - 6'   ~ 5.5  ,
                            month == '7 - 8'   ~ 7.5   , 
                            .default = as.numeric(month))) -> NHM_filterred2

### Identifying Broods
NHM_filterred2 %>%
  left_join( select(BroodTable, Include ,
                    `Taxon Name`= `Taxonomic species` ,
                    Univoltine, FirstMonthOf2ndBrood),
             by = 'Taxon Name')%>%
  filter( !(!Univoltine&is.na(month ) )) %>% # remove bivoltine with no month info
  mutate(Brood = ifelse( Univoltine ,'Univoltine',
                         ifelse( month < FirstMonthOf2ndBrood,
                                 'Brood1',
                                 'Brood2'))) -> NHM_filterred_Broods

### For each sample, identifying season of temp records to use

NHM_filterred_Broods%>%
  left_join(AllBroodTable_Seasons, 
            by = c('Taxon Name','Brood')) %>%
  mutate( T_DEVELOP= case_match( TempDEVELOP ,
                                 'SPRING'        ~  paste0('SEASON_', year, '-04-16'),
                                 'SUMMER'        ~  paste0('SEASON_', year, '-07-16'),
                                 'SUMMER_before' ~  paste0('SEASON_', year-1, '-07-16')),
          T_LAY= case_match( TempLAY,
                             'SPRING'        ~  paste0('SEASON_', year, '-04-16'),
                             'SUMMER'        ~  paste0('SEASON_', year, '-07-16'),
                             'SUMMER_before' ~  paste0('SEASON_', year-1, '-07-16')))%>%
  filter( Include == 'TRUE')%>%
  filter( decimalLatitude> 49.5)-> FilteredSizeAndLoc_Seasons
```


### Loading Temps 

```{r}
FilePathTo.nc<- "../data/dap.ceda.ac.uk/badc/ukmo-hadobs/data/insitu/MOHC/HadOBS/HadUK-Grid/v1.3.0.ceda/25km/tas/seas/v20240514/"
Files<- list.files(FilePathTo.nc,full.names = TRUE)[-1] ## dropping 'index'
List_Rasters<- map(Files, rast)
## ChangeNamesToTimes
List_Rasters2<- map(List_Rasters, function(RS){names(RS)<-  paste0('SEASON_', time(RS));return(RS)})
## Collapse the list
Collated_SeasonalTemps<- rast(List_Rasters2)
```

```{r}
UKGRID_COORDS<-sf_project(pts=  select(FilteredSizeAndLoc_Seasons,decimalLongitude,decimalLatitude ),
                          from ='epsg:4326', 
                          to ='+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +a=6377563.396 +rf=299.324961266495 +units=m +no_defs')

### Checking matches are all possible
Seasons_ToMatch<- sort(unique(c(FilteredSizeAndLoc_Seasons$T_DEVELOP,
                                FilteredSizeAndLoc_Seasons$T_LAY )))
all(Seasons_ToMatch %in% names(Collated_SeasonalTemps))

## Extracting all years out into a table of specimen x seasons
RecordTemps<- terra::extract(x = Collated_SeasonalTemps,
                             y=UKGRID_COORDS)
## Checking alignment is good
plot(Collated_SeasonalTemps[[1]] )
points( UKGRID_COORDS) 

## There will be some points that just miss a box. For them can setsearch_radius = 100, but not with multilayers
## Need to pull out RecordTemps

SeasonTemp_Dev<- c(NA, nrow( RecordTemps))
SeasonTemp_Lay<- c(NA, nrow( RecordTemps))


for( i in 1:nrow(RecordTemps)){
  ## dumb but not that slow way to get numbers out of matrix
  SeasonTemp_Dev[i] <- RecordTemps[i ,FilteredSizeAndLoc_Seasons$T_DEVELOP[i] ]
  SeasonTemp_Lay[i] <- RecordTemps[i ,FilteredSizeAndLoc_Seasons$T_LAY[i]     ]
}

FilteredSizeAndLoc_Seasons$SeasonTT_Develop <- SeasonTemp_Dev
FilteredSizeAndLoc_Seasons$SeasonTT_Lay <- SeasonTemp_Lay
```


```{R eval = FALSE}
FilteredSizeAndLoc_Seasons %>%
  ggplot()+
  geom_point(aes( x = decimalLongitude, y = decimalLatitude,
                  col = is.na(SeasonTT_Lay)))  ## Just weird coastal ones.

## Fixing these up:Can use nearest neagbour

which(is.na(FilteredSizeAndLoc_Seasons$SeasonTT_Develop)) -> WeirdCoastal_Numbs

for(i in WeirdCoastal_Numbs){   ## takes a little while to run (~5 minutes)
  print(i)
  ## Development Period
  SeasonLayer = FilteredSizeAndLoc_Seasons$T_DEVELOP[i] 
  yyy<- terra::extract(x = subset(Collated_SeasonalTemps,SeasonLayer) ,
                       y=vect(UKGRID_COORDS)[i],
                       search_radius=100000) ## up to 100km to nearest box
  SeasonTemp_Dev[i] <- yyy[1,2]
  
  ## Laying Period
  SeasonLayer = FilteredSizeAndLoc_Seasons$T_LAY[i] 
  zzz<- terra::extract(x = subset(Collated_SeasonalTemps,SeasonLayer) ,
                       y=vect(UKGRID_COORDS)[i],
                       search_radius=100000) ## up to 100km to nearest box
  SeasonTemp_Lay[i] <- zzz[1,2]
}

## Replacing
FilteredSizeAndLoc_Seasons$SeasonTT_Develop <- SeasonTemp_Dev
FilteredSizeAndLoc_Seasons$SeasonTT_Lay <- SeasonTemp_Lay

### Checking all good now:
FilteredSizeAndLoc_Seasons %>%
  ggplot()+
  geom_point(aes( x = decimalLongitude, y = decimalLatitude,
                  col = is.na(SeasonTT_Develop)))  ## Just weird coastal ones. Also Channel islands?

write_csv(FilteredSizeAndLoc_Seasons, 
          '../data/FilteredSizeAndLoc_Seasons_DRYRUN.csv')

```

```{r}
FilteredSizeAndLoc_Seasons <- read_csv(  '../data/FilteredSizeAndLoc_Seasons_DRYRUN.csv')
```


# Oxford Measurements

## Which have matching names?

```{r}
OUNHM_orig <- readxl::read_excel("../data/Project_data_copynew.xlsx", sheet = "Sheet1") 
UA_temps <- read_csv( '../data/UA_Seasonal_Temps_Long.csv') #  
VC_temps <- read_csv('../data/VC_Seasonal_Temps_Long.csv')

Ox_Locs_To_Match <- unique(OUNHM_orig$Location)  

VC_names <- sort(unique(VC_temps$VCNAME))
NotInUA<- Ox_Locs_To_Match[!(Ox_Locs_To_Match %in% unique(UA_temps$CTYUA23NM))]

NotInUA
NotInUA[NotInUA %in% VC_names]
NameNotInEither <- NotInUA[!(NotInUA %in% VC_names)]

NameNotInEither

OUNHM_orig %>%
  filter( Location %in% NameNotInEither)
```

## Which locations in both?

```{r}

NamesInBoth<-VC_names[VC_names %in% unique(UA_temps$CTYUA23NM)]
### If in both, will use VC names as probably better match to older records


VC_temps %>%
  mutate(LocationType = 'VC') %>%
  select( Location= VCNAME, LocationType,
          MeanTemp , 
          Date = MidDate) -> VC_Temps_simple
  
UA_temps %>%
  mutate(LocationType = 'VC') %>%
  select( Location= CTYUA23NM, LocationType,
          MeanTemp , 
          Date = MidDate) %>% 
  filter(!(Location %in% NamesInBoth)) -> UA_Temps_simple

  
  bind_rows(VC_Temps_simple, UA_Temps_simple)-> All_Loc_Temps


```


## Joining

```{R}
OUNHM_orig



OUNHM_orig %>% 
  mutate(  Month_num = match(str_to_sentence(Month),# reading mess of different capitalisations
                             month.name)) %>%
  filter(Year > 1885) %>%
  select(Species, Sex,
         month = Month_num,
         year =Year, 
         Location ,
         Wing_Length,
         Abdomen_Width,
         Abdomen_Length, 
         Indiv_Num = Individual) -> OxNHM_Simp

### Identifying Broods

##################  Needs names to do joining!!!

# OxNHM_Simp %>%
# ### For each sample, identifying season of temp records to use
#   left_join(AllBroodTable_Seasons, 
#             by = c('Taxon Name','Brood')) %>%
#   mutate( T_DEVELOP= case_match( TempDEVELOP ,
#                                  'SPRING'        ~  paste0('SEASON_', year, '-04-16'),
#                                  'SUMMER'        ~  paste0('SEASON_', year, '-07-16'),
#                                  'SUMMER_before' ~  paste0('SEASON_', year-1, '-07-16')),
#           T_LAY= case_match( TempLAY,
#                              'SPRING'        ~  paste0('SEASON_', year, '-04-16'),
#                              'SUMMER'        ~  paste0('SEASON_', year, '-07-16'),
#                              'SUMMER_before' ~  paste0('SEASON_', year-1, '-07-16')))%>%
#   filter( Include == 'TRUE')%>%
#   left_join()




```








