---
title: "Quick Spin through NHM data"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Using data from: Wilson et al https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13844


Sourced from: 
https://zenodo.org/records/5759759

and

https://zenodo.org/records/6066070


```{r}
library( tidyverse)
library(readxl)

Project_data_copynew <- read_excel("../data/Project_data_copynew.xlsx", sheet = "Sheet1") 

```


## Loading in measurments

```{r}

## Different groups saved in sperate sheets

# Sheets<- excel_sheets('../data/NHM/Wilson_et_al_MEE_data.xlsx')
# 
# map_df()
# 
# read_xlsx('../data/NHM/Wilson_et_al_MEE_data.xlsx', 
#             )

WilsonSH3<- read_xlsx('../data/NHM/Wilson_et_al_MEE_data.xlsx', sheet = 3)

Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')

```

### Downloading locations

Gordon L J Paterson; Sara Albuquerque; Vladimir Blagoderov; Stephen J Brooks et al. (2016). iCollections [Data set]. Natural History Museum. https://doi.org/10.5519/0038559

 *DOI has been created for this download request: https://doi.org/10.5519/qd.tqng8ri6 (this may take a few hours to become active). Please ensure you reference this DOI when citing these data.
For more information, follow the DOI link.*


```{r eval = FALSE}
Sizes<- read_csv('../data/NHM/iCollections_Nov_2021_butterflies.csv')

NHM_data <- read_csv("C:/Users/pemb4504/Downloads/7ba5ee8575924a906078531b9212aec880045725/resource.csv")

colnames(NHM_data)

NHM_data %>%
  select( "_id","basisOfRecord","catalogNumber",
          "class","collectionCode",
          "continent","coordinateUncertaintyInMeters", "country",                 
          "created","day","decimalLatitude","decimalLongitude" ,            
          "determinationFiledAs","determinationNames",  
          "family", "gbifID"    ,                   
          "gbifIssue", "genus",
          "higherClassification","month",      
          "order","project" ,                     
          "recordedBy" ,                 
          "scientificName",               
          "verbatimLatitude","verbatimLongitude","year" )      -> NHM_data_firstcut  


ifelse(sapply(NHM_data_firstcut, function(x)all(is.na(x))) == TRUE, "Y","N")

NHM_data_firstcut %>%
  mutate( `specimen number` = parse_number(catalogNumber))  %>%
  select( catalogNumber,  `specimen number`,determinationNames ,
          decimalLatitude,decimalLongitude, year )-> Just_Names_Loc_and_Numbers

write_csv(Just_Names_Loc_and_Numbers, '../data/Just_Names_Loc_and_Numbers.csv')
```


```{r}
Just_Names_Loc_and_Numbers <- read_csv( '../data/Just_Names_Loc_and_Numbers.csv')

 left_join(Sizes, Just_Names_Loc_and_Numbers) -> Joined_sizes_and_Locs

Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
  ggplot(aes( x = decimalLongitude, y = decimalLatitude, col = year))+
   geom_point()


Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
  count(`Taxon Name`, sort = TRUE)

Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
  filter(  `Taxon Name`  == 'Lysandra coridon') %>%
  ggplot(aes( x = decimalLongitude, y = decimalLatitude, col = year))+
   geom_point()

Joined_sizes_and_Locs%>%
  filter(!is.na( decimalLatitude ))%>%
   ggplot(aes( y = `wing_span (mm)`, x = year))+
   geom_point()

```


# Temperature effects

```{r}
WilsonSH3 %>% count(Species )

```







