---
title: "OUNHM cleaning 4"
author: "Elsa Heywood"
date: "2025-06-20"
output: html_document
---
## Adding the shapefiles to get midpoint data for each location and then combining that with temperature, brood, and name data.
# Load libraries
```{r}
library(sf)
library(dplyr)
library(tidyverse)
library(readxl)
```

# Read in shapefile data for vice counties and unitary authorities
```{r}
counties <- st_read("../data/vice-counties-master/vice-counties-master/12-mile/County_12mile_region.shp")
print(counties)

unitary <- st_read("../data/Counties_and_Unitary_Authorities_May_2023_UK_BGC_9070630924157476989/CTYUA_MAY_2023_UK_BGC.shp")
print(unitary)

```

# First extract midpoints, convert to LatLong for unitary authorities and clean
```{r}
# Get centroids of each unitary authority polygon
unitary_centroids <- st_centroid(unitary)

# Transform the centroids to WGS84 coordinate system (EPSG:4326, which is lat/lon)
unitary_centroids <- st_transform(unitary_centroids, crs = 4326)

# Extract coordinates into a data frame
centroid_coords <- st_coordinates(unitary_centroids)

# Combine coordinates with attribute data (like the name of each unitary authority)
unitary_midpoints <- unitary_centroids %>%
  st_drop_geometry() %>%
  bind_cols(as.data.frame(centroid_coords)) %>%
  rename(longitude = X, latitude = Y)

#clean columns
unitary_midpoints_clean <- unitary_midpoints %>%
  select(CTYUA23CD, CTYUA23NM, longitude, latitude)
```

# View and save
```{r}
# View the result
head(unitary_midpoints_clean)

write_csv(unitary_midpoints_clean, "../data/unitary_authority_midpoints.csv")

```

# Same for vice counties
```{r}
##midpoints for counties
st_crs(counties)

# Get centroids
county_centroids <- st_centroid(counties)

# Transform to WGS84 (lat/lon)
county_centroids <- st_transform(county_centroids, crs = 4326)

# Extract coordinates
coords <- st_coordinates(county_centroids)

# Combine with attribute data
county_midpoints <- county_centroids %>%
  st_drop_geometry() %>%
  bind_cols(as.data.frame(coords)) %>%
  rename(longitude = X, latitude = Y)

#clean columns
county_midpoints_clean <- county_midpoints %>%
  select(VCNUMBER, VCNAME, longitude, latitude)
```

# View and save 
```{r}
# View result
head(county_midpoints_clean)

write_csv(county_midpoints_clean, "../data/county_midpoints.csv")

```

# Combine the unitary authority and vice county midpoint data, to do this I have to add their LocationType
```{r}
# Add type to unitary authorities
unitary_midpoints_clean <- unitary_midpoints_clean %>%
  mutate(Name = CTYUA23NM,
         LocationType = "UA") %>%
  select(Name, LocationType, latitude, longitude)

# Add type to counties
county_midpoints_clean <- county_midpoints_clean %>%
  mutate(Name = VCNAME,
         LocationType = "VC") %>%
  select(Name, LocationType, latitude, longitude)
# Combine
combined_locations <- bind_rows(unitary_midpoints_clean, county_midpoints_clean)
head(combined_locations)
combined_locations %>% filter(LocationType == "VC") %>% head()

write_csv(combined_locations, "../data/combined_midpoints.csv")
```

## Now combine this with OUNHM_data_temps and that with BroodTable and Names_table
# Read in data
```{r}
OUNHM_data_temps <- read_csv("../data/OUNHM_data_temps.csv")
BroodTable <- read_xlsx("../data/butterfly brood table.xlsx")
Names_table <- read_xlsx("../data/names_table.xlsx")

```

# Clean and combine brood tables
```{r}
Brood_table2 <- BroodTable %>%
  left_join(Names_table, by = 'NHM name') %>%
  mutate(FirstMonthBrood2 = as.numeric(FirstMonthBrood2)) %>%
  filter(Include == 'TRUE') %>%
  select(`Taxon name`, Brood1Use, Brood2Use, FirstMonthBrood2, Univoltine)

# Remove duplicate rows from Brood_table2
Brood_table2_unique <- Brood_table2 %>%
  distinct()

Brood_table_long <- Brood_table2_unique %>%
  select(-FirstMonthBrood2, -Univoltine) %>%
  pivot_longer(
    names_to = 'Brood',
    values_to = 'TempToUse',
    cols = Brood1Use:Brood2Use
  ) %>%
  filter(TempToUse != 'NA') %>%
  mutate(Brood = parse_number(Brood))

Brood_table_long_unique <- Brood_table_long %>%
  group_by(`Taxon name`, Brood) %>%
  slice(1) %>%
  ungroup()
```

# Add all tables including location_midpoints
```{r}
OUNHM_for_models <- OUNHM_data_temps %>%
  left_join(
    combined_locations %>%
      select(Name, LocationType, midpointLongitude = longitude, midpointLatitude = latitude),
    by = c("Location" = "Name", "LocationType" = "LocationType")
  ) %>%
  # Join Brood_table2 for FirstMonthBrood2 and Univoltine
  left_join(
    select(Brood_table2_unique, `Taxon name`, FirstMonthBrood2, Univoltine),
    by = 'Taxon name'
  ) %>%
  mutate(Brood = ifelse(Univoltine, 1,
                        ifelse(Month < FirstMonthBrood2, 1, 2))) %>%
  # Join brood temp table
  left_join(Brood_table_long_unique, by = c('Taxon name', 'Brood')) %>%
  select(`Taxon name`, midpointLatitude, midpointLongitude, Year,
         Wing_Length, Brood, TempToUse,
         SPRING_temp, SUMMER_temp, SUMMER_before_temp) %>%
  mutate(Temp = case_when(
    TempToUse == 'Spring' ~ SPRING_temp,
    TempToUse == 'Summer' ~ SUMMER_temp,
    TempToUse == 'Summer_before' ~ SUMMER_before_temp,
    TRUE ~ NA_real_
  )) %>%
  select(-ends_with('_temp')) %>%
  distinct()  # Remove duplicates
```

# View and save
```{r}
OUNHM_for_models
write_csv(OUNHM_for_models, "../data/OUNHM_for_models.csv")
```

