---
title: "StatsFitting2_AcrossSpecies"
author: "Chris Terry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

Main document for making main text results

With Sex embedded as a predictor of size in each species model

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(posterior)
library(forcats)
library(bayesplot)
library(cowplot)
```

```{r}
## Summaries of posteriors of each model

PS_NHM<- read_csv('../statsresults/NHM_Posterior_Summaries_M1.csv')
PS_OUM<- read_csv('../statsresults/OUM_Posterior_Summaries_M1.csv')
```

```{r}
# Predictors

# Specialism 
specialism <- read_csv( '../data/specialism_with_nhm.csv')%>%
  mutate(Label = case_match(Specialism, 
                            'Habitat and foodplant generalist' ~ 'Habitat generalist\n&\nfoodplant generalist',
                            'Habitat and foodplant specialist'~ 'Habitat specialist\n&\nfoodplant specialist',
                            'Habitat generalist and foodplant specialist'~ 'Habitat generalist\n&\nfoodplant specialist',
                            'Habitat specialist and foodplant generalist'~  'Habitat specialist \n&\nfoodplant generalist'))


read_csv('../data/brood_table_CTupdates.csv') %>%
  filter(Include == 'TRUE' )%>%
  select(Sp = `Taxon name` ,Family )-> SpFamily



## Butterfly Red List https://resjournals.onlinelibrary.wiley.com/doi/10.1111/icad.12582

UKRedList <- read_csv('../data/ButterflyRedlistUK.csv')

```

# NHM

## Figure For paper

### Prepping data

```{r}

PS_NHM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  distinct(Specialism, Sp, Family ) %>%
  arrange(Specialism, Sp) -> OrderTable


ORDER_f2 <-pluck(OrderTable,'Sp')

OrderTable$Family_Cols <- c( 'darkgreen',"blue",  'orange',"gold", 'darkgrey', '#800020')[factor(pluck(OrderTable,'Family'))]


## Colour legend

Legend_plot<- OrderTable %>%
  ggplot(aes(x = Family))+
  geom_point( aes( col = Family), y = 1, size = 3, shape =15)+
  scale_color_manual(values =  c( 'darkgreen',"blue",  'orange',
                                  "gold", 'darkgrey', '#800020'))+
  theme(legend.position = 'bottom')

PS_NHM %>%
  mutate(Sp  = factor(Sp, levels = ORDER_f2)) %>% 
  mutate( q5 = if_else(variable== 'b_Latitude' & Sp =='Melitaea cinxia',
                       -Inf, q5)) %>% 
  mutate( q16.7 = if_else(variable== 'b_Latitude' & Sp =='Melitaea cinxia',
                          -Inf, q16.7)) %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  filter( variable %in% c( 'b_Temp',
                           'b_YearRescale',
                           'b_Sexmale',
                           'b_Latitude') )%>%
  mutate(Response = case_match(variable, 
                               'b_Temp'~ 'b. Temperature',
                               'b_YearRescale'~'c. Year',
                               'b_Sexmale'~'d. Male Offset',
                               'b_Latitude'~'a. Latitude') ) -> DataForMainNHMPlot


```

### Main Text Figure
```{r}
DataForMainNHMPlot%>%
  mutate(Redlist = ifelse( is.na(`GB Red List category`),
                           'Unlisted', 'Listed')) %>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean))+
  geom_errorbarh(aes( xmin = q5, xmax = q95), linewidth = 1, height = 0, col = 'grey')+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3), linewidth = 2, height = 0, col = 'grey')+
  geom_point(shape = 4)+
  geom_vline(xintercept = 0)+
  facet_grid(Redlist~Response, scales = 'free', space = "free_y")+
  ggtitle('((NHM datasets, Species-level coefficients, Monthly, Facetted by Redlist)')+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient')

ggsave('../figs/NHM_Monthly_MainSpResults_byRedlist.png', width = 10, height = 7, bg = 'white')

```

### SI Figures

```{r}
DataForMainNHMPlot%>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean))+
  geom_errorbarh(aes( xmin = q5, xmax = q95), linewidth = 1, height = 0, col = 'grey')+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3), linewidth = 2, height = 0, col = 'grey')+
  geom_point(shape = 4)+
  geom_vline(xintercept = 0)+
  facet_grid(Label~Response, scales = 'free', space = "free_y")+
  ggtitle('((NHM datasets, Species-level coefficients, Monthly, Facetted by Generalism cat))')+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient')

ggsave('../figs/NHM_Monthly_MainSpResults_by4wayGeneralism.png', width = 10, height = 7, bg = 'white')

DataForMainNHMPlot%>%
  mutate(Generalist = ifelse(Specialism== 'Habitat and foodplant generalist',
                             'GENERALIST', 'SPECIALIST')) %>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean))+
  geom_errorbarh(aes( xmin = q5, xmax = q95), linewidth = 1, height = 0, col = 'grey')+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3), linewidth = 2, height = 0, col = 'grey')+
  geom_point(shape = 4)+
  geom_vline(xintercept = 0)+
  facet_grid(Generalist~Response, scales = 'free', space = "free_y")+
  ggtitle('((NHM datasets, Species-level coefficients, Monthly, Facetted by Generalism cat))')+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient')

ggsave('../figs/NHM_Monthly_MainSpResults_by2wayGeneralism.png', width = 10, height = 7, bg = 'white')


DataForMainNHMPlot%>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean))+
  geom_errorbarh(aes( xmin = q5, xmax = q95), linewidth = 1, height = 0, col = 'grey')+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3), linewidth = 2, height = 0, col = 'grey')+
  geom_point(shape = 4)+
  geom_vline(xintercept = 0)+
  facet_grid(Family~Response, scales = 'free', space = "free_y")+
  ggtitle('((NHM datasets, Species-level coefficients, Monthly, Facetted by Family)')+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient')

ggsave('../figs/NHM_Monthly_MainSpResults_byFamily.png', width = 10, height = 7, bg = 'white')


PS_NHM %>%
  filter( variable == 'b_YearRescale' ) %>%
  brm( mean|mi(sd)  ~  1 ,
       data = .) -> AverageYearEffect

AverageYearEffect %>% posterior_summary()

# OrderTable$Family_Cols
# 
# plot_grid( SpNHMResultsPlot,
#            get_legend(Legend_plot),
#            rel_heights = c(10,1) , ncol = 1)

```

## Tests of predictors

### Prepping data

```{r}
NHM_for_fitting <- read_csv('../data/NHM_for_models_monthly.csv')
NHM_for_fitting %>%
  group_by(Sp)%>%
  summarise(MeanRightWing = signif(mean(Right_wing),4) ,
            SD_RightWing = signif(sd(Right_wing),4),
            CV_RightWing = signif(sd(Right_wing)/mean(Right_wing),4)) -> SpeciesSizeProps

PS_NHM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  left_join(SpeciesSizeProps) %>%
  left_join(SpFamily) %>%
  mutate(RedListListed = ifelse(is.na( `GB Red List category`),
                                'Unlisted',
                                'Listed')) %>% 
  filter( variable %in% c( 'b_Temp',
                           'b_YearRescale') )%>%
  mutate(ConfidentTrend =  sign(q5)== sign(q95),
         TrendSign = ifelse(ConfidentTrend, sign(mean), 0)) %>%
  mutate( Generalist = Specialism == 'Habitat and foodplant generalist') %>%
  select(variable,Sp, Specialism , Family, RedListListed , TrendSign, Generalist,
         mean, sd, MeanRightWing,SD_RightWing ,CV_RightWing) -> Trends

```

### Correlations Between Three Predictors

```{R}
Trends %>%
  filter( variable == 'b_Temp') %>%
  mutate( RedListListed = RedListListed== 'Listed') %>%
  select(RedListListed, MeanRightWing, Generalist ) %>%
  as.matrix() %>%cor %>%
  corrplot::corrplot.mixed(lower.col = 'black') 

Trends %>%
  ggplot(aes( x = RedListListed, y = MeanRightWing))+
  geom_boxplot()+
  geom_point()+
  geom_label(aes( label = Sp))+
  scale_y_log10()


Trends %>% 
  filter( variable == 'b_Temp') %>%
  group_by(RedListListed)%>%
  count( Generalist) ->redlist_generalist_table

redlist_generalist_table

redlist_generalist_table%>%
  pluck('n') %>%
  matrix(.,ncol = 2) %>%
  chisq.test()

cor(Trends$Generalist, Trends$RedListListed=='Listed' )  # same as phi

```

### Predictors by Family

```{R}
Trends %>%
  filter( variable == 'b_Temp') %>%
  select( Sp, Family, RedListListed, TrendSign, Generalist, MeanRightWing)->x

FamilyCorrelates_a <- ggplot(x, aes( x = Family, fill = RedListListed))+
  geom_bar()+theme_bw()+theme(legend.position="bottom")
FamilyCorrelates_b <- ggplot(x, aes( x = Family, y = MeanRightWing))+
  geom_boxplot()+geom_point()+theme_bw()+theme(legend.position="bottom")
FamilyCorrelates_c <- ggplot(x, aes( x = Family, fill = Generalist))+
  geom_bar()+theme_bw()+theme(legend.position="bottom")
FamilyCorrelates_d <- ggplot(x, aes( x = Family, fill = factor(TrendSign)))+
  geom_bar()+theme_bw()+theme(legend.position="bottom")

plot_grid( FamilyCorrelates_a,
           FamilyCorrelates_b,
           FamilyCorrelates_c,
           FamilyCorrelates_d, ncol = 1)

```

## Predictors of Trends

TrendSign = just taking the three-way split of coefficients - positive, indeterminate, negative. 

### Red List Status 

```{r}

Trends %>%
  ggplot(aes( x = RedListListed, y= mean))+
  geom_boxplot()+
  geom_point(aes(col = factor(TrendSign)))+
  facet_wrap(~variable, scales = 'free_y')+
  geom_text(aes(label = Sp), size = 2 ,hjust=-0.15 )+
  theme_bw()

Trends %>%
  filter( variable == 'b_Temp' ) %>%
  ggplot(aes( x = RedListListed, y= mean))+
  geom_boxplot()+
  geom_point(aes(col = factor(TrendSign)))+
  geom_text(aes(label = Sp), size = 2 ,hjust=-0.15 )+
  theme_bw()+
  ylab('Temperature Coefficent')

Trends %>% 
  group_by(variable, RedListListed)%>%
  count( TrendSign) %>%
  pivot_wider(names_from = RedListListed,
              id_cols =c(variable,TrendSign),
              values_from = n ) %>%
  t %>%
  knitr::kable()

chisq.test(matrix(c(10,5,
                    4,12),2,2, byrow = TRUE)) 

chisq.test(matrix(c(10,5,
                    12,9,
                    4,12),3,2, byrow = TRUE))

PS_NHM %>%
  filter( variable == 'b_Temp' ) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  mutate(RedListListed = ifelse(is.na( `GB Red List category`),
                                'Unlisted', 'Listed')) %>%
  brm( mean|mi(sd)  ~  RedListListed ,
       data = .) -> RedList_TempEffect_Model

RedList_TempEffect_Model %>% posterior_summary()


PS_NHM %>%
  filter( variable == 'b_YearRescale' ) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  mutate(RedListListed = ifelse(is.na( `GB Red List category`),
                                'Unlisted', 'Listed')) %>%
  brm( mean|mi(sd)  ~  RedListListed ,
       data = .) -> RedList_YearEffect_Model

RedList_YearEffect_Model %>% posterior_summary()
```

### Generalism

```{r}

Trends %>%
  ggplot(aes( x = Generalist, y= mean))+
  geom_boxplot()+
  geom_point(aes(col = factor(TrendSign)))+
  facet_wrap(~variable, scales = 'free_y')+
  geom_text(aes(label = Sp), size = 2 ,hjust=-0.15 )

Trends %>%
  group_by(variable, Generalist)%>%
  count( TrendSign) %>%
  pivot_wider(names_from = Generalist,
              id_cols =c(variable,TrendSign),
              values_from = n ,
              values_fill = 0) %>%
  t %>%
  knitr::kable()

## Temperature pattern not significant however 0s are treated:
chisq.test(matrix(c(7,8,
                    13,3),2,2, byrow = TRUE))

chisq.test(matrix(c(7,8,
                    14,7,
                    13,3),3,2, byrow = TRUE))

## Year Pattern
chisq.test(matrix(c(5,0,16,11),2,2, byrow = TRUE))

chisq.test(matrix(c(5,0,
                    13,7,
                    16,11),3,2, byrow = TRUE))

Trends %>%
  filter( variable == 'b_YearRescale' ) %>%
  brm( mean|mi(sd)  ~   Generalist     ,
       data = .) -> Generalist_YearEffect_Model

Generalist_YearEffect_Model %>% posterior_summary()


Trends %>%
  filter( variable == 'b_Temp' ) %>%
  brm( mean|mi(sd)  ~   Generalist     ,
       data = .) -> Generalist_TempEffect_Model

Generalist_TempEffect_Model %>% posterior_summary()


```

### Size

```{r}
Trends %>%
  ggplot(aes( x = factor(TrendSign), y = MeanRightWing))+
  geom_boxplot()+
  geom_point()+
  facet_wrap(~variable)

PS_NHM %>%
  filter( variable == 'b_YearRescale' ) %>%
  left_join(SpeciesSizeProps) %>%
  brm( mean|mi(sd)  ~  MeanRightWing ,
       data = .) -> Size_YearEffect_Model

Size_YearEffect_Model %>% posterior_summary() %>% as.data.frame()-> SizeEffectSummary
SizeEffectSummary

mcmc_hist(Size_YearEffect_Model,
          pars = 'b_MeanRightWing')+
  xlab('Slope of Average Size vs bYear Coefficient,\naccounting for uncertainty')+
  theme_bw()-> PosteriorSizevsYear

Trends %>%
  filter( variable == 'b_YearRescale' ) %>%
  ggplot(aes( x = MeanRightWing, y= mean))+
  geom_abline(intercept = SizeEffectSummary$Estimate[1],
              slope = SizeEffectSummary$Estimate[2])+
  geom_point()+
  geom_text(aes(label = Sp), size = 2)+
  theme_bw()+
  xlab('Butterfly Average Size,\n(Mean Right Wing)')+
  ylab('Species_level Posterior Mean of\nProportional Size Response to Year') -> SizevsTrendWithYear

plot_grid(SizevsTrendWithYear, PosteriorSizevsYear, rel_widths = c(2,1))

ggsave('../figs/SizeEffectOnTemproralTrend.png', height = 4, width = 8)


### Testing if effect can be explained as well just by family
PS_NHM %>%
  filter( variable == 'b_YearRescale' ) %>%
  left_join(SpeciesSizeProps) %>%
  left_join(SpFamily) %>%
  brm( mean|mi(sd)  ~  Family ,
       data = .) -> FamilyGroup_YearEffect_Model 

MS <- add_criterion(Size_YearEffect_Model, criterion = 'loo')
MF <- add_criterion(FamilyGroup_YearEffect_Model, criterion = 'loo')

loo_compare(MS, MF)  # Can't differentate 
```

### Family

Excluding families with only one member

```{r}
Trends %>%
  #filter( !Sp %in%  c('Papilio machaon', 'Hamearis lucina' )) %>%
  group_by(variable, Family)%>%
  count( TrendSign) %>%
  pivot_wider(names_from = Family,
              id_cols =c(variable,TrendSign),
              values_from = n ,
              values_fill = 0) ->FamilyDataForChiSq

FamilyDataForChiSq%>%
  t %>%
  knitr::kable()

Trends %>%
  ggplot(aes( x = Family, y = mean))+
  geom_boxplot()+
  geom_point()+
  facet_wrap(~variable, scales = 'free_y')


FamilyDataForChiSq %>%
  filter( variable == 'b_Temp' ) %>%
  ungroup()%>%
  select(Hesperiidae,Lycaenidae,Nymphalidae,Pieridae ) %>%
  as.matrix() %>%
  chisq.test()

```




# Quantiative Estimates of Responses


## Predictors of b_Temp 

```{r eval = FALSE}
PS_NHM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name'))%>%
  filter(Specialism !='Habitat specialist and foodplant generalist') %>% ## trying excluding the two HS/FG
  filter( variable =='b_Temp' ) -> bTemp_pred_NHM

bTemp_pred_NHM %>%
  brm( median|mi(sd)  ~  Specialism,
       data = ., iter = 1000)%>%
  add_criterion( criterion = c("loo")) -> bTemp_model_NHM

bTemp_pred_NHM %>%
  brm( median|mi(sd)  ~  1,
       data = ., iter = 1000)%>%
  add_criterion(criterion = c("loo")) -> bTemp_model_NHM_NoSpec

save( bTemp_model_NHM, file = '../TopLevelModelFits/bTemp_model_NHM')
save( bTemp_model_NHM_NoSpec,file =  '../TopLevelModelFits/bTemp_model_NHM_NoSpec')
```

```{r}
load('../TopLevelModelFits/bTemp_model_NHM')
load('../TopLevelModelFits/bTemp_model_NHM_NoSpec')

loo_compare(bTemp_model_NHM,
            bTemp_model_NHM_NoSpec ) ### It is worth including specialism? Can't identify a clear signal

plot(conditional_effects(bTemp_model_NHM, resp = "median", rug = TRUE), ask = FALSE)
```


## Predictors of b_year

```{r eval = FALSE}
PS_NHM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name'))%>%
  filter(Specialism !='Habitat specialist and foodplant generalist') %>% ## trying excluding the two HS/FG
  filter( variable =='b_YearRescale' ) -> bYear_pred_NHM

bYear_pred_NHM %>%
  brm( median|mi(sd)  ~  Specialism,
       data = ., iter = 1000)%>%
  add_criterion( criterion = c("loo")) -> bYear_model_NHM

bYear_pred_NHM %>%
  brm( median|mi(sd)  ~  1,
       data = ., iter = 1000)%>%
  add_criterion(criterion = c("loo")) ->bYear_model_NHM_NoSpec


save( bYear_model_NHM, file = '../TopLevelModelFits/bYear_model_NHM')
save( bYear_model_NHM_NoSpec, file = '../TopLevelModelFits/bYear_model_NHM_NoSpec')
```

```{r}
load('../TopLevelModelFits/bYear_model_NHM')
load('../TopLevelModelFits/bYear_model_NHM_NoSpec')

loo_compare(bYear_model_NHM,
            bYear_model_NHM_NoSpec ) ### It is worth including specialism? Can't identify a clear signal

plot(conditional_effects(bYear_model_NHM, resp = "median", rug = TRUE), ask = FALSE)
```

## Sample Size vs confidence

```{r}
NHM_ForFitting <- read_csv(  '../data/NHM_for_models_monthly.csv')

PS_NHM %>%
  filter( variable %in% c('b_YearRescale' , 'b_Temp'   , 'b_Latitude')) %>%
  left_join(count(NHM_ForFitting, Sp)) %>%
  mutate(ConfidentTrend =  sign(q5)== sign(q95)) %>%
  ggplot(aes( x = n, y = sd))+
  geom_point(aes(col = ConfidentTrend))+
  scale_x_sqrt()+
  facet_wrap(~variable, scales = 'free')+
  ylab('SD of posterior of fitted coefficents')+
  xlab('Sample size (NB Square-root scale)')+
  ggtitle('Relationship between data size for each species and uncertainty in coefficent')

ggsave('../figs/Corr_sample_size_trends_NHM.png', width = 9, height = 3)

PS_NHM %>%
  filter( variable %in% c('b_YearRescale' , 'b_Temp'   , 'b_Latitude')) %>%
  left_join(count(NHM_ForFitting, Sp)) %>%
  mutate(ConfidentTrend =  sign(q5)== sign(q95)) %>%
  ggplot(aes( x = ConfidentTrend, y = sqrt(n)))+
  geom_boxplot(aes(fill = ConfidentTrend))+
  geom_point()+
  facet_wrap(~variable, scales = 'free')

```

# OUNHM

## Figure for paper main text

```{r}
# ORDER_f3 <- PS_OUM %>%
#   left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
#   distinct(Specialism, Sp ) %>%
#   arrange(Specialism, Sp) %>%pluck('Sp')

PS_OUM %>%
  mutate(Sp  = factor(Sp, levels = ORDER_f3)) %>% 
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  filter( variable %in% c( 'b_Temp',
                           'b_YearRescale',
                           'b_Sexmale',
                           'b_Latitude') )%>%
  mutate(Response = case_match(variable, 
                               'b_Temp'~ 'b. Temperature',
                               'b_YearRescale'~'c. Year',
                               'b_Sexmale'~'d. Male Offset',
                               'b_Latitude'~'a. Latitude') )%>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean))+
  geom_errorbarh(aes( xmin = q5, xmax = q95), linewidth = 1, height = 0, col = 'grey')+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3), linewidth = 2, height = 0, col = 'grey')+
  geom_point(shape = 4)+
  geom_vline(xintercept = 0)+
  facet_grid(~Response, scales = 'free', space = "free_y")+
  ggtitle('((OUNHM datasets, Species-level coefficients, Monthly))',
          'Colours are currrently random, but will be family')+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient')


##  theme(axis.text.y = element_text(colour = c("gold", "blue", 'darkgreen')))

ggsave('../figs/OUNHM_Monthly_MainSpResults.png', width = 10, height = 5, bg = 'white')
```

## Predictors of b_Temp 

```{r eval = FALSE}
PS_OUM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  left_join(SpeciesSizeProps)%>%
  mutate(RedListListed = ifelse(is.na( `GB Red List category`),
                                'Unlisted',
                                'Listed')) %>%
  filter( variable =='b_Temp' )-> bTemp_model_OUM


bTemp_model_OUM %>%
  ggplot(aes( x = MeanRightWing , y =mean     ))+
  geom_point()+
  geom_smooth(method = 'lm')


bTemp_model_OUM %>%
  brm( median|mi(sd)  ~  Specialism,  data = .)%>%
  add_criterion(criterion = c("loo"))-> bTemp_model_OUM

bTemp_model_OUM %>%
  brm( median|mi(sd)  ~  1,  data = .)%>%
  add_criterion(criterion = c("loo")) -> bTemp_model_OUM_NoSpec


bTemp_model_OUM %>%
  brm( median|mi(sd)  ~ RedListListed ,  data = .)%>%
  add_criterion(criterion = c("loo"))-> RedList_bTemp_model_OUM

RedList_bTemp_model_OUM %>% summarise_draws()



save( bTemp_model_OUM, file = '../TopLevelModelFits/bTemp_model_OUM')
save( bTemp_model_OUM_NoSpec, file = '../TopLevelModelFits/bTemp_model_OUM_NoSpec')
save( RedList_bTemp_model_OUM, file = '../TopLevelModelFits/RedList_bTemp_model_OUM')

```

```{r}
load('../TopLevelModelFits/bTemp_model_OUM')
load('../TopLevelModelFits/bTemp_model_OUM_NoSpec')

loo_compare(bTemp_model_OUM,bTemp_model_OUM_NoSpec ) ### It is worth including specialism

bTemp_model_OUM

plot(conditional_effects(bTemp_model_OUM , resp = "median", rug = TRUE), ask = FALSE)
```

## Predictors of b_Year

```{r eval = FALSE}

PS_OUM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  left_join(SpeciesSizeProps)%>%
  filter( variable =='b_YearRescale' )-> bYear_model_OUM

bYear_pred_OUM %>%
  brm( median|mi(sd)  ~  Specialism,  data = .)%>%
  add_criterion(criterion = c("loo"))-> bYear_model_OUM

bYear_pred_OUM %>%
  brm( median|mi(sd)  ~  1,  data = .)%>%
  add_criterion(criterion = c("loo")) -> bYear_model_OUM_NoSpec



save( bYear_model_OUM, file = '../TopLevelModelFits/bYear_model_OUM')
save( bYear_model_OUM_NoSpec, file = '../TopLevelModelFits/bYear_model_OUM_NoSpec')
```

```{r}
load('../TopLevelModelFits/bYear_model_OUM')
load('../TopLevelModelFits/bYear_model_OUM_NoSpec')

loo_compare(bYear_model_OUM,
            bYear_model_OUM_NoSpec ) ### It is worth including specialism

bYear_model_OUM

bYear_model_OUM_NoSpec %>% summarise_draws()



plot(conditional_effects(bTemp_model_OUM , resp = "median", rug = TRUE), ask = FALSE)
```

# Comparison to Wilson fits

### Species-level Plots

```{r}

PS_WIL <- read_csv('../statsresults/WIL_Posterior_Summaries_X2.csv')


bind_rows(PS_NHM %>%
            filter( variable %in% c( 'b_Temp',
                                     'b_Latitude',
                                     'b_YearRescale') ) %>%
            mutate( Dataset= 'NHM'), 
          PS_WIL %>%
            filter( variable %in% c( 'b_Wilson_LateTemp',
                                     #    'b_Latitude',
                                     'b_YearRescale') )%>%
            mutate( Dataset= 'Wilson'),
          PS_OUM %>%
            filter( variable %in% c( 'b_Temp',
                                     'b_Latitude',
                                     'b_YearRescale') )%>%
            mutate( Dataset= 'OUNHM')) %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) -> Joined_coefficients

ORDER <- Joined_coefficients %>%
  filter(variable=='b_Temp', Dataset =='NHM' ) %>%
  arrange(mean) %>%pluck('Sp')


Joined_coefficients%>%
  ggplot( aes( y =  Sp , 
               x = mean, 
               col = Dataset))+
  geom_point()+
  geom_errorbarh(aes( xmin = q5, xmax = q95))+
  geom_vline(xintercept = 0)+
  scale_y_discrete(limits = ORDER)+
  facet_grid(.~variable, scales = 'free')+
  ggtitle('Species-level coefficients')


ggsave(filename = '../figs/compare_coefs_across_three_datasetsM1.png', width = 10, height =6)



Joined_coefficients%>%
  filter(variable%in% c( 'b_Wilson_LateTemp')) %>%
  ggplot( aes( y =  Sp , 
               x = mean, 
               col = Dataset))+
  geom_point()+
  geom_errorbarh(aes( xmin = q5, xmax = q95))+
  geom_vline(xintercept = 0)+
  #scale_y_discrete(limits = ORDER)+
  facet_grid(Specialism~variable, scales = 'free')+
  ggtitle('Species-level coefficients')



## widen dataset to compare coefficents across datasets 

Joined_coefficients %>%
  select(  variable,
           mean, q5, q95, Sp, Dataset, Specialism) %>%
  pivot_wider( id_cols = c('variable', 'Specialism','Sp' ),
               names_from = 'Dataset',
               values_from = c( 'mean','q5','q95' )) %>%
  filter( !is.na(mean_OUNHM )) %>%
  ggplot( aes( y =  mean_NHM , 
               x = mean_OUNHM))+
  geom_point(aes(col = Specialism))+
  geom_errorbarh(aes(col = Specialism,xmin = q5_OUNHM, xmax = q95_OUNHM))+
  geom_errorbar(aes(col = Specialism,ymin = q5_NHM, ymax = q95_NHM))+
  geom_abline(intercept = 0, slope =1)+
  geom_vline(xintercept = 0)+
  geom_hline(yintercept = 0)+
  facet_wrap(~variable, scales = 'free')+
  #coord_fixed()+
  theme(legend.position = 'bottom')+
  ggtitle('Correlation in Species-level coefficients')

ggsave(filename = '../figs/compare_coefs_across_datasets.png', width = 10, height =6)

```

# Session Info

```{r}
sessionInfo()
```